---
title: "3. Ecosystem compatibility: TraMineR, tna, and Nestimate"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{3. Ecosystem compatibility: TraMineR, tna, and Nestimate}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(transitiontrees)
has_tna  <- requireNamespace("tna",       quietly = TRUE)
has_nest <- requireNamespace("Nestimate", quietly = TRUE)
```

`transitiontrees` is built to slot into the wider `Dynalytics`  
ecosystem (`tna`, `Nestimate`, `cograph`) as well as `TraMineR`.
You do **not** have to re-export or re-format your data: `context_tree()` 
accepts a fitted state-sequence or transition/network object **directly**, 
auto-detected by its S3 class. It reads the sequences the object already 
carries and fits the suffix tree on exactly those, over the same state set.

This vignette shows the three hand-offs on one shared dataset and confirms
they agree. The `tna` and `Nestimate` sections only run if those packages are
installed; the `TraMineR` state-sequence section runs on the bundled data
with no extra package.

## The shared dataset

The bundled `engagement` object is a `TraMineR` state-sequence (`stslist`):
1000 weekly engagement sequences over three states.

```{r data}
data(engagement)
class(engagement)
dim(engagement)
```

## Route A -- a state-sequence (`stslist`) object

Hand the `stslist` straight to `context_tree()`.

```{r stslist-naive}
tree_sts_raw <- context_tree(engagement, max_depth = 2L, min_count = 5L)
tree_sts_raw$alphabet
```

Note the alphabet: a `TraMineR` sequence object encodes gaps/missing with a
**void code** (here `%`), and taken literally that void looks like a fourth
state. The clean fix is to pass the object's declared `alphabet` so the void
is treated as end-of-sequence, not a state:

```{r stslist-clean}
tree_sts <- context_tree(engagement, max_depth = 2L, min_count = 5L,
                        alphabet = attr(engagement, "alphabet"))
tree_sts$alphabet
n_nodes(tree_sts)
```

That is the tree to use. The `tna` and `Nestimate` objects below carry the
clean label set internally, so they decode the void for you automatically --
the two routes converge on this same tree.

## Route B -- a `tna` transition-network object

A `tna` model carries its underlying sequences in its `$data` slot.
`context_tree()` reads them and decodes through the model's label set.

```{r tna, eval = has_tna}
library(tna)
model_tna <- tna(engagement)
class(model_tna)

tree_tna <- context_tree(model_tna, max_depth = 2L, min_count = 5L)
tree_tna
```

## Route C -- a `Nestimate` network object

`Nestimate::build_tna()` (and the other `build_*()` constructors) return a
`netobject` that likewise carries the sequence frame and a `$nodes` label
table. Same hand-off:

```{r nestimate, eval = has_nest}
library(Nestimate)
model_nest <- build_tna(engagement)
class(model_nest)

tree_nest <- context_tree(model_nest, max_depth = 2L, min_count = 5L)
tree_nest
```

## They agree

Because all three objects wrap the *same* sequences, the fitted trees are
identical -- same alphabet, same nodes, same observation count.

```{r agree, eval = has_tna && has_nest}
identical(tree_sts$nodes, tree_tna$nodes)
identical(tree_tna$nodes, tree_nest$nodes)

data.frame(
  route   = c("stslist", "tna", "Nestimate"),
  n_nodes = c(n_nodes(tree_sts), n_nodes(tree_tna), n_nodes(tree_nest)),
  nobs    = c(model_fit(tree_sts)$nobs, model_fit(tree_tna)$nobs,
              model_fit(tree_nest)$nobs))
```

The tree shares one symbol space with the source model, so the whole
pathway API -- `common_pathways()`, `divergent_pathways()`,
`bootstrap_pathways()`, the plots -- composes onto an already-estimated
network with no conversion step.

## The boundary: sequences, never aggregated transitions

The hand-off works only when the object actually **carries its sequences**. A
*pure graph* projection -- nodes, edges and weights with the raw sequences
nulled out (an aggregated transition network) -- is **rejected**, on purpose:
the original sequences cannot be recovered from edge weights, so fabricating
them would be silently wrong. The same invariant rejects a bare numeric
transition matrix.

```{r pure-graph, eval = has_nest, error = TRUE}
graph_only <- build_tna(engagement)
graph_only$data <- NULL          # strip the stored sequences
context_tree(graph_only)         # -> informative error, not a fabricated fit
```

`transitiontrees` fits on raw sequences. If you have only an aggregated
network, go back to the sequence data it was built from.

## Group objects

The grouped constructors compose too. `context_tree()` recognises a
`group_tna` / `netobject_group` (a named list of per-group models) and fits
one tree per group, returning a `transitiontrees_group` that
`prune_tree()`, `compare_trees()`, and `compare_groups()` consume directly --
see the *Advanced analysis* vignette for the group workflow.
