---
title: "1. Getting started: basic analysis and trajectory trees"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    highlight: tango
    df_print: paged
vignette: >
  %\VignetteIndexEntry{1. Getting started: basic analysis and trajectory trees}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 8, fig.height = 5, out.width = "100%")
library(transitiontrees)
```

`transitiontrees` fits a **variable-depth prediction suffix tree** to
categorical sequence data and reports it as a tidy, pathway-centric set of
tables and plots. A fixed-order Markov chain assumes memory is the *same
length everywhere*; a variable-order tree lets the **data decide, per
context, how much history matters**. This first vignette walks the core
workflow end to end and finishes with the two **trajectory trees** that
draw the sequences forward in time.

The other vignettes go further: *Complete analysis case* reads one dataset
all the way through, *Ecosystem compatibility* shows the `TraMineR` / `tna` /
`Nestimate` hand-off, *Advanced analysis* covers tuning, bootstrapping and
comparison, and *Visualization* tours every plot.

## 1. Fit

`context_tree()` accepts a wide character matrix or data.frame, a list of
character vectors, a state-sequence (`stslist`) object, or a long event
log. We start with the bundled `trajectories` matrix (138 learners x 15
time-steps, three engagement states; trailing `NA`s mark dropouts).

```{r fit}
data(trajectories)
dim(trajectories)

tree <- context_tree(trajectories, max_depth = 3L, min_count = 5L)
tree
```

`max_depth` caps how long a history (context) may be; `min_count` is the
minimum number of times a context must occur to earn its own node.

A **long event log** is reshaped internally -- just name the columns:

```{r long}
data(group_regulation_long)
head(group_regulation_long)
tree_long <- context_tree(group_regulation_long,
                          actor = "Actor", time = "Time", action = "Action",
                          max_depth = 2L, min_count = 5L)
n_nodes(tree_long)
```

## 2. Inspect the fit

```{r inspect}
summary(tree)
model_fit(tree)   # logLik, df, nobs, AIC, BIC, perplexity
```

Perplexity is the effective number of equally likely next states; it sits
below the uniform baseline (the alphabet size, here 3) when history is
informative.

## 3. The pathway tables

Every accessor returns a plain `data.frame` in one canonical schema, so the
views join cleanly. Pathways read left-to-right oldest-to-newest
(`A -> B -> C`); the root context is shown as `(start)`.

 - `common_pathways`:Returns the top n pathways by occurrence count – the trajectories the data actually contains many copies of.
 - `divergent_pathways`: Returns the top n pathways by Kullback-Leibler divergence from their (k-1)-suffix. These are the pathways whose extended history adds the most predictive information over the shorter one. Pathways whose most likely next state actually flips between orders are marked in the changes_prediction column.
 - `sharp_pathways`: Returns the top n pathways by predictive sharpness – the probability mass on their modal next state. High values indicate strongly deterministic continuations; low values indicate ambiguous next-state distributions.


```{r tables}
common_pathways(tree, top = 6)      # by frequency
divergent_pathways(tree, top = 6)   # by divergence from the shorter history
sharp_pathways(tree, top = 6)       # by how peaked the next-state prediction is
```

`changes_prediction = TRUE` flags a context whose single most likely next
state differs from its parent's -- the histories where memory overturns the
corpus-wide default. The lesson the tables teach together: **common is not
the same as informative**. The most frequent pathways are the backbone of
the corpus; the divergent ones carry the insight.

## 4. Prune to the reliable tree

A context can survive fitting yet not earn its depth. `prune_tree()`
collapses contexts whose extra history is not a significant gain over their
parent (default: a likelihood-ratio G-squared test).

```{r prune}
pruned <- prune_tree(tree, criterion = "G2", alpha = 0.05)
pruned
```

The pruned tree's banner reports its node count and the criterion used --
compare it to the unpruned `tree` printed in section 1 to see how much the
G-squared test removed.

## 5. Predict

```{r predict}
predict(pruned, c("Active", "Active"), type = "class")          # most likely next
round(predict(pruned, c("Active", "Active"), type = "prob"), 3) # full distribution
```

When an exact context is missing from the tree, prediction backs off to the
longest matching suffix -- the property that makes a *variable*-order model
robust: it never refuses to predict, it just uses as much history as it has
evidence for.

## 6. A first tree plot

Just `plot()` the tree. The default is a horizontal layout: node size is the
context count, the colour is the most-recent state, and the predicted next
state sits under each node.

```{r plot, fig.width = 14, fig.height = 8}
plot(pruned)
```

`plot()` also takes `style = "dendrogram"`, `"icicle"`, or `"interactive"`
for the same tree in other layouts -- the *Visualization* vignette tours all
four.

## 7. Trajectory trees: where sequences go, and how predictably

The context tree reads *backwards* -- a node is a suffix, the most-recent
state. The same sequences can be drawn *forwards* as a **trajectory tree**:
start at the left, every path is a run of states unfolding in time. Forward
trajectories are most informative on a richer alphabet, so we switch to the
bundled `ai_long` log -- one row per AI-prompting move (eight move types:
`Execute`, `Investigate`, `Plan`, ...), with a session id. `context_tree()`
reads it directly.

```{r traj-fit}
data(ai_long)
tree_ai   <- context_tree(ai_long, actor = "project", session = "session_id",
                         action = "code", max_depth = 3L, min_count = 10L)
pruned_ai <- prune_tree(tree_ai)
tree_ai
```

`plot_trajectories()` draws the forward prefix tree and colours the one tree
two ways. These plots use `ggforce` (a suggested package), so the chunks are
guarded.

### By frequency -- how many sequences walk each path

```{r traj-frequency, fig.width = 11, fig.height = 7, eval = requireNamespace("ggforce", quietly = TRUE)}
plot_trajectories(tree_ai, measure = "frequency", min_count = 20L)
```

Node fill and edge width both scale to the number of sessions on each path,
so the thick, dark branches are the prompting routines most projects
actually follow -- the corpus's highways from the opening move outward.

### By predictability -- how confidently the model calls each step

```{r traj-predictability, fig.width = 11, fig.height = 7, eval = requireNamespace("ggforce", quietly = TRUE)}
plot_trajectories(pruned_ai, measure = "predictability", min_count = 20L)
```

Same nodes and edges, but each edge is now coloured by `P(state | history)`
from the model. Reading the two side by side separates **traffic** from
**predictability**: an edge that is wide (frequency) but pale
(predictability) is a *decision point* -- many sessions reach it, but the
next move is genuinely open. Those forks are where behaviour is decided
rather than executed.

## Where to go next

| You want to... | See vignette |
|---|---|
| Read one dataset all the way through | *Complete analysis case* |
| Feed in a `TraMineR` / `tna` / `Nestimate` object | *Ecosystem compatibility* |
| Tune, bootstrap, and compare cohorts | *Advanced analysis* |
| Tour every plot style | *Visualization* |
