---
title: "5. Visualization: every plot, and how to read it"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    highlight: tango
    df_print: paged
vignette: >
  %\VignetteIndexEntry{5. Visualization: every plot, and how to read it}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 8, fig.height = 5, out.width = "100%")
has_ggraph  <- requireNamespace("ggraph", quietly = TRUE) &&
               requireNamespace("tidygraph", quietly = TRUE)
has_visnet  <- requireNamespace("visNetwork", quietly = TRUE)
has_ggforce <- requireNamespace("ggforce", quietly = TRUE)
```

`transitiontrees` ships a pure-`ggplot2` visual layer (no JavaScript
dependency in the core) plus two optional richer renderers. Every plot
returns a standard object you can theme, save, or further modify. This
vignette tours them all and reads each one.

A shared convention across the static tree styles: **node size = context
count**, **node fill = the most-recent state of the pathway**, and **edge
thickness = the volume of sequences flowing down that branch**.

## Setup

Every plot below is drawn from the same fitted, pruned tree on the bundled
`trajectories` data (138 learners, three engagement states). We fit it once
here and reuse it throughout.

```{r setup}
library(transitiontrees)
data(trajectories)
set.seed(1)

tree   <- context_tree(trajectories, max_depth = 3L, min_count = 5L)
pruned <- prune_tree(tree, criterion = "G2", alpha = 0.05)
pruned
```

## 1. The fitted tree, four ways

### Horizontal phylogram (default)

Root on the left, depth rightward; every leaf is labelled with its full
arrow-form pathway and the predicted next state. This is the style for a
paper when you need to cite specific pathways inline.

```{r horizontal, fig.width = 14, fig.height = 8}
plot(pruned, style = "horizontal")
```

`point_size_range` and `edge_size_range` exaggerate or compress the size
dynamic range -- useful for slides where the count contrast must read from
the back of the room. The encodings are unchanged; only the scales differ.

```{r horizontal-sized, fig.width = 14, fig.height = 8}
plot(pruned, style = "horizontal",
     point_size_range = c(3, 12), edge_size_range = c(0.4, 3.5))
```

### Radial dendrogram

The same tree wrapped into a circle: the eye goes to the thick central
branches (the corpus highways) versus the thin outer twigs (contexts pruning
kept on evidence, not volume).

```{r dendrogram, fig.height = 6}
plot(pruned, style = "dendrogram")
```

### Icicle / sunburst (needs `ggraph` + `tidygraph`)

A space-filling partition: arc angular width is proportional to count, so a
dominant state visually swallows the ring -- an honest depiction of class
imbalance.

```{r icicle, fig.height = 6, eval = has_ggraph}
plot(pruned, style = "icicle")
```

### Interactive collapsible tree (needs `visNetwork`)

A draggable, zoomable widget carrying the same encodings, with a hover
tooltip for each node's full pathway, count, and distribution. The
exploration tool: collapse the dominant spine and the rare informative
branches become legible.

```{r interactive, fig.height = 6, eval = has_visnet}
plot(pruned, style = "interactive")
```

## 2. Pathway-centric plots

These complement the tree by ranking *pathways* rather than drawing topology.

### Next-state heatmap

Each row is a context, each column a next state, each cell
`P(next | context)`, modal cell bold; a `>` prefix marks a context whose
modal next state flips versus its shorter parent. Sorting the **same** data
two ways is the single best "common vs informative" figure:

```{r heatmap-count, fig.height = 5.5}
plot_pathways(pruned, top = 12, sort_by = "count")        # the highways
```

```{r heatmap-div, fig.height = 5.5}
plot_pathways(pruned, top = 12, sort_by = "divergence")   # the informative ones
```

Sorted by count the bright cells stack on the most frequent next state;
sorted by divergence they move off it. That lateral shift is the thesis in
one comparison.

### Divergence lollipop

Per-context KL from the shorter parent, ranked, with orange points marking
modal-flip contexts -- the histories that genuinely *change* the prediction.
`min_count` removes small-sample mirages.

```{r divergence, fig.height = 5}
plot_divergence(pruned, top = 12, min_count = 5)
```

### Per-context distributions

The full next-state distribution for each context as small multiples --
peaked panels are near-settled continuations, flat panels are the decision
points where history does not resolve the next state.

```{r distributions, fig.height = 5.5}
plot_distributions(pruned, top = 6)
```

## 3. Diagnostic plots

### How much memory does one pathway need?

`plot_pruning()` walks a pathway's suffix chain -- the full context, then the
same context with its oldest move dropped, down to the root -- and marks
which contexts the pruning test keeps (solid) versus drops (faded). It
answers, for that one pathway, how far back history actually has to reach.

```{r pruning, fig.width = 9, fig.height = 4.5}
plot_pruning(tree, "Active -> Active -> Average")
```

### Predictive quality

`plot_predictive()` scores sequences against the fitted tree three ways. For
this tour we score the bundled `trajectories` themselves; in a real
evaluation pass genuinely held-out sequences (the *Advanced analysis*
vignette shows the cross-validated route).

`type = "logloss"` -- per-position surprise in bits against position; below
the uniform ceiling is structure the model exploited:

```{r predictive-logloss, fig.height = 4.5}
plot_predictive(pruned, trajectories, type = "logloss")
```

`type = "ecdf"` -- the distribution of the probability assigned to the state
that actually occurred; steep steps reveal calibration plateaus (e.g. a mass
of three-way-open branch points):

```{r predictive-ecdf, fig.height = 4.5}
plot_predictive(pruned, trajectories, type = "ecdf")
```

A third type, `type = "position"`, traces each individual sequence's
confidence move-by-move (one grey line per sequence). It is a per-sequence
view that only reads cleanly for a handful of sequences, so it is omitted
here; reach for it when you want to inspect a few specific trajectories
rather than the corpus as a whole.

## 4. Forward trajectory trees (needs `ggforce`)

The context tree reads backward; `plot_trajectories()` draws the same
sequences forward in time. Colour by **frequency** (how many sequences walk
each path) or by **predictability** (`P(state | history)` from the model).
Read together they separate traffic from predictability -- a wide-but-pale
edge is a high-traffic decision point.

Forward trajectories show their structure best on a richer alphabet, so this
section uses the bundled `ai_long` log (eight AI-prompting move types) rather
than the three-state engagement data above.

```{r traj-fit}
data(ai_long)
tree_ai   <- context_tree(ai_long, actor = "project", session = "session_id",
                         action = "code", max_depth = 3L, min_count = 10L)
pruned_ai <- prune_tree(tree_ai)
```

```{r traj-freq, fig.width = 11, fig.height = 7, eval = has_ggforce}
plot_trajectories(tree_ai, measure = "frequency", min_count = 20L)
```

```{r traj-pred, fig.width = 11, fig.height = 7, eval = has_ggforce}
plot_trajectories(pruned_ai, measure = "predictability", min_count = 20L)
```

## 5. Inferential plots

### Bootstrap forest plot

Each pathway's 95% bootstrap interval on G-squared against the chi-square
critical value (dashed line); colour encodes the trust quadrant. A bar
entirely to the right is reproducibly informative.

```{r boot-plot, fig.height = 5.5}
boot <- bootstrap_pathways(pruned, iter = 200L, seed = 1L)
plot(boot)
```

### Per-pathway resample distributions

```{r boot-resamples, fig.height = 4.5}
plot_pathway_resamples(boot, stat = "divergence", top = 6)
```

### Cohort comparison: permutation null

We name an external group column (`Achiever`) on the bundled
`group_regulation_long` log; `context_tree(group = )` fits one tree per
cohort, and `compare_trees()` consumes the group directly.

```{r compare-plot, fig.height = 4.5}
data(group_regulation_long)
grp_reg <- context_tree(group_regulation_long,
                       actor = "Actor", time = "Time", action = "Action",
                       group = "Achiever", max_depth = 2L, min_count = 10L)
cmp <- compare_trees(prune_tree(grp_reg), iter = 199L, seed = 1L)
plot(cmp)
```

The observed distance (orange line) sits in the right tail of the
label-shuffled null (grey) -- the visual form of the permutation p-value.

### Tuning surface

```{r tune-plot, fig.height = 5}
tg <- tune_tree(trajectories, max_depth = 1L:4L, folds = 5L, seed = 1L)
plot(tg)
```

A flat-then-rising perplexity curve is the picture of a short-memory process;
the orange star marks the cross-validated winner.

### Group difference map

`plot_difference()` draws the per-context residual map for the same
`group =`-fitted tree -- where two cohorts resolve the same history toward
different next states. `depth = 1L` keeps the map to the single-state
contexts so the rows stay legible (a deep tree has too many contexts to label).

```{r difference, fig.height = 5}
plot_difference(grp_reg, depth = 1L)
```

## Recap

| Goal | Function |
|---|---|
| The tree | `plot(style = c("horizontal", "dendrogram", "icicle", "interactive"))` |
| Rank pathways | `plot_pathways()`, `plot_divergence()`, `plot_distributions()` |
| Memory of one pathway | `plot_pruning()` |
| Held-out quality | `plot_predictive(type = c("logloss", "ecdf", "position"))` |
| Forward trajectories | `plot_trajectories(measure = c("frequency", "predictability"))` |
| Reliability | `plot(<bootstrap>)`, `plot_pathway_resamples()` |
| Comparison | `plot(<comparison>)`, `plot_difference()` |
| Tuning | `plot(<tune>)` |
