---
title: "transitiontrees::bootstrap_pathways() — demonstration"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    theme: flatly
    df_print: paged
    code_folding: show
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE,
  fig.width = 9, fig.height = 5, dpi = 110, out.width = "100%"
)
devtools::load_all(here::here())
```

## What `bootstrap_pathways()` reports

Methodologically built on Saqr, Tikka & López-Pernas (2025)
(doi:10.1145/3706468.3706513), extending an edge-level bootstrap
framework to variable-depth pathways.

The bootstrap operates on **raw counts only** — no `nmin` filter,
no smoothing inside the loop, no extra parameters. Each iteration
resamples whole sequences, aggregates counts per depth, and reads
each original-tree pathway's count vector directly.

Two complementary measures per pathway:

- **`stability_rate`** — fraction of resamples where the chosen
  `stat` (default `count`) lands in
  `[0.75 × observed, 1.25 × observed]`. **Value reproducibility.**
  A pathway whose count drops to zero in a resample fails the band
  test automatically — the same mechanism an edge-level bootstrap
  uses for edge weights.

- **`informative_rate`** — fraction of resamples where the
  pathway's empirical likelihood-ratio statistic against its parent
  context exceeds the chi-square critical value at `alpha =
  0.05`. **Reproducible significant divergence from the
  shorter-history baseline.** The bootstrap analogue of
  `prune(criterion = "G2")`.

A pathway is `stable` when `stability_rate ≥ stability_threshold`
(default 0.75) and `informative` when `informative_rate ≥
informative_threshold` (default 0.95). The summary sorts
**stable + informative** pathways first.

---

## Validation: strong, weak, random

Three Markov regimes from a small inline first-order Markov
generator, same sample size (100 sequences × 14 length, alphabet
`{A, B, C}`):

```{r sim}
P_strong <- rbind(c(0.05, 0.90, 0.05),
                  c(0.05, 0.05, 0.90),
                  c(0.90, 0.05, 0.05))
P_weak   <- rbind(c(0.30, 0.50, 0.20),
                  c(0.20, 0.30, 0.50),
                  c(0.50, 0.20, 0.30))
P_random <- matrix(1 / 3, 3L, 3L)
rownames(P_strong) <- colnames(P_strong) <- c("A", "B", "C")
rownames(P_weak)   <- colnames(P_weak)   <- c("A", "B", "C")
rownames(P_random) <- colnames(P_random) <- c("A", "B", "C")
init <- c(A = 1/3, B = 1/3, C = 1/3)

# Inline first-order Markov sequence generator: returns a
# character matrix (n_sequences rows × seq_length columns).
simulate_markov <- function(n_sequences, seq_length, trans_matrix,
                            init_probs, seed) {
  set.seed(seed)
  states <- colnames(trans_matrix)
  step <- function(prev, ignore) {
    sample(states, 1L, prob = trans_matrix[prev, ])
  }
  one_seq <- function(i) {
    first <- sample(states, 1L, prob = init_probs)
    Reduce(step, seq_len(seq_length - 1L), accumulate = TRUE,
           init = first)
  }
  do.call(rbind, lapply(seq_len(n_sequences), one_seq))
}

run_scenario <- function(P, seed) {
  seqs <- simulate_markov(
    n_sequences = 100L, seq_length = 14L,
    trans_matrix = P, init_probs = init, seed = seed)
  tree <- context_tree(seqs, max_depth = 2L, min_count = 5L)
  bootstrap_pathways(tree, iter = 300L, seed = 1L)
}

b_strong <- run_scenario(P_strong, seed = 1L)
b_weak   <- run_scenario(P_weak,   seed = 2L)
b_random <- run_scenario(P_random, seed = 3L)
```

### Print: STRONG (modal probability 0.90)

```{r print-strong}
b_strong
```

Three depth-1 pathways (A, B, C) have G² ~600 with tight CIs —
their conditional distribution differs dramatically from the
marginal, so `informative_rate = 1`. Depth-2 pathways have G² near
zero because the data is 1st-order Markov: conditional on the most
recent state, prior history adds nothing. The bootstrap correctly
flags depth-1 as informative and depth-2 as not — the latter would
be the wrong choice in a longer-memory tree.

### Print: WEAK (modal probability 0.50)

```{r print-weak}
b_weak
```

Same structure, weaker signal. Depth-1 G² ~50–60: still well
above the chi-square critical value (5.99), so depth-1 is
reproducibly informative. Depth-2 G² ~1–4 sits below the
critical value most of the time, so depth-2 is reproducibly not
informative — exactly right for 1st-order Markov.

### Print: RANDOM (uniform 1/3)

```{r print-random}
b_random
```

All counts are reproducible (`stable = TRUE` everywhere — counts
of uniform random transitions reproduce just fine), but **zero
pathways are informative** at the strict 0.95 threshold. The
chi-square test is hovering at finite-sample chance levels (G²
~5–14, near or just above the critical value), and the bootstrap
correctly says it doesn't *reproducibly* clear the threshold.

### Headline numbers across regimes

```{r headline}
cmp <- function(b, label) {
  s <- b$summary
  data.frame(
    scenario      = label,
    n_pathways    = nrow(s),
    n_stable      = sum(s$stable, na.rm = TRUE),
    n_informative = sum(s$informative, na.rm = TRUE),
    n_both        = sum(s$stable & s$informative, na.rm = TRUE),
    median_G2     = round(median(s$mean_G2, na.rm = TRUE), 2),
    mean_G2       = round(mean(s$mean_G2, na.rm = TRUE), 2),
    p95_G2        = round(quantile(s$mean_G2, 0.95,
                                    na.rm = TRUE), 2)
  )
}
rbind(
  cmp(b_strong, "STRONG (modal=0.90)"),
  cmp(b_weak,   "WEAK   (modal=0.50)"),
  cmp(b_random, "RANDOM (uniform)")
)
```

`mean_G2` orders cleanly across regimes (151 / 16 / 4) and
`n_informative` orders the same way (3 / 3 / 0) at the strict
threshold. `n_stable` is high in all three because counts are
reproducible regardless of underlying structure — that's
stability's job; informativeness is the test that distinguishes
real signal.

### Forest plots

```{r forest-strong, fig.height = 4}
plot(b_strong)
```

```{r forest-weak, fig.height = 4}
plot(b_weak)
```

```{r forest-random, fig.height = 4}
plot(b_random, min_stability = 0)
```

The dashed line marks the chi-square critical value. STRONG's
depth-1 pathways sit far to the right of it; WEAK's still clear it
comfortably; RANDOM's straddle or fall below it.

### Resample distributions of G²

```{r resamples-strong, fig.height = 3.6}
plot_pathway_resamples(b_strong, stat = "G2", top = 6L)
```

```{r resamples-random, fig.height = 3.6}
plot_pathway_resamples(b_random, stat = "G2", top = 6L)
```

STRONG's depth-1 G² distributions are tight at 500–700; depth-2
distributions sit near zero. RANDOM shows everything piled up
near the chi-square critical value with wide spread — exactly what
finite-sample noise produces.

---

## Real data: `trajectories`

```{r real-data}
data(trajectories)
dim(trajectories)
tree <- context_tree(trajectories, max_depth = 3L, min_count = 5L)
boot <- bootstrap_pathways(tree, iter = 500L, seed = 1L)
boot
```

```{r real-plot, fig.height = 6}
plot(boot)
```

```{r real-resamples, fig.height = 4}
plot_pathway_resamples(boot, stat = "G2", top = 6L)
```

```{r real-summary}
summary(boot)
```

---

## Design notes

- **Bootstrap operates on raw counts.** No `nmin` filter, no
  smoothing inside the loop, no extra parameters. Pathways tracked
  are exactly the ones in the original tree; their values come
  from per-depth count aggregation.
- **`stable` answers "does the value reproduce?"** — the
  edge-level bootstrap framing. With enough data, counts reproduce regardless of
  whether the underlying structure is informative.
- **`informative` answers "does the LR test reproducibly clear the
  threshold?"** — the bootstrap analogue of
  `prune(criterion = "G2")`. This is the test that distinguishes
  real signal from finite-sample noise.
- **Use both flags together.** A pathway worth interpreting is
  one where `stable AND informative`. The summary sorts those
  first.
- **No `prune()` inside the resample loop.** Pruning is a
  downstream filter on the bootstrap output, not part of the fit
  replay.
- **Performance.** Per-sequence `(context, next_state)`
  pair-index vectors are precomputed once per depth (a
  transition fast-path). Bootstrap iterations are one
  `unlist`+`tabulate` per depth followed by per-pathway count
  lookups, with no `data.frame` construction inside the loop.

---

```{r session, echo = FALSE}
sessionInfo()
```
