---
title: "Loading data into transitiontrees"
subtitle: "Every input shape context_tree() accepts"
author: "transitiontrees"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_depth: 2
    toc_float:
      collapsed: false
    df_print: paged
    highlight: tango
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      comment = "#>")
suppressPackageStartupMessages(devtools::load_all(here::here()))
```

`transitiontrees` is interoperable with fitted transition/network
objects: a model carrying its underlying sequences can be passed
**directly** to `context_tree()`, which recovers that sequence data
(auto-detected by the object's S3 class) and estimates the suffix tree on
it, with no re-formatting, re-import, or conversion of any kind — the tree
is defined over exactly the same state set as the source model.
transitiontrees also follows the **standard long→sequence reshaping
vocabulary**, so reading raw, long-format data (one row per event) is
straightforward: `actor` identifies the unit a sequence belongs to, `time`
(or an explicit `order`) sequences the events, `action` names the state
observed at each event, and sessions are delimited either by an explicit
`session` column or by a `time_threshold` gap. These are supplied to
`prepare_input()`, or to `context_tree()` directly through the matching
arguments. Finally, for data that is already organised as sequences,
`context_tree()` accepts a wide table (one row per case, columns advancing
through time), a ragged list of sequences, and a state-sequence `stslist`
object. The sections below illustrate each of these routes on example
datasets bundled with the package, so the vignette runs without any
external data.

# 1. A wide matrix or data.frame

The simplest shape: rows are sequences, columns are time-steps, cells are
states, trailing `NA`s mark the end of a sequence.

```{r wide}
data(trajectories)
head(trajectories)
context_tree(trajectories)
```

> A character **matrix** or **data.frame** is accepted as-is. (Numeric
> matrices are rejected — they mean "transition matrix" elsewhere in the
> ecosystem — so cast to character first.)

# 2. A list of ragged sequences

When sequences have different lengths, a list of character vectors avoids
padding.

```{r list}
seqs <- list(c("A", "B", "A", "C"),
             c("B", "B"),
             c("C", "A", "A"))
context_tree(seqs, max_depth = 2L, min_count = 1L)
```

> Each list element is one sequence; missing cells (`NA` or `""`) are
> dropped, so a sequence keeps only its observed states.

# 3. A state-sequence object (`stslist`)

If your data is already a state-sequence `stslist` object, hand it
straight over — the alphabet and any weights are honoured.

```{r stslist}
data(engagement)         # an stslist of engagement states
context_tree(engagement, max_depth = 2L)
```

# 4. Long format (one row per event)

The usual real-world shape is *long*: each row is a single event, with
columns for who it belongs to (`actor`), when it happened (`time`), and
what it was (`action`).

## 4a. In one call

Name the state column with `action =` and `context_tree()` reshapes the
data for you — grouping by `actor`, ordering by `time`, and splitting
each actor's events into **sessions** wherever the time gap is large.

```{r long-onecall}
data(group_regulation_long)
head(group_regulation_long)
context_tree(group_regulation_long, actor = "Actor", time = "Time",
             action = "Action", max_depth = 3L)
```

## 4b. Or reshape first with `prepare_input()`

`prepare_input()` does the same reshaping and returns the wide frame, so
you can inspect it before fitting. The two routes are identical.

```{r long-twostep}
wide <- prepare_input(group_regulation_long, actor = "Actor",
                      time = "Time", action = "Action")
head(wide)
context_tree(wide, max_depth = 3L)
```

> `prepare_input()` is a standard long→sequence reshaper: a new session
> starts when the gap to the previous event exceeds `time_threshold`
> seconds (default 900). The result is the wide sequence frame from §1.

# 5. Long-format options: sessions, order, Unix time

`prepare_input()` (and the `context_tree()` arguments that forward to it)
handle the common variations. Here on `ai_long`, whose `timestamp` is
Unix time and which carries an explicit `session_id`.

```{r prep-options}
data(ai_long)
## (i) timestamp gaps -> sessions (Unix time is auto-detected)
gap <- prepare_input(ai_long, actor = "project", time = "timestamp",
                     action = "code", time_threshold = 600)
dim(gap)

## (ii) use an explicit session column instead of time gaps
sess <- prepare_input(ai_long, actor = "project", session = "session_id",
                      action = "code")
dim(sess)

## (iii) no timestamp: order events with an order column
ord <- prepare_input(ai_long, actor = "project",
                     order = "order_in_session", action = "code")
dim(ord)
```

> `time_threshold` controls how aggressively a `time` column is split
> into sessions; `session =` takes sessions from a column directly; and
> with neither, `order =` (or row order) sequences each actor.

# 6. From a fitted transition/network object

`context_tree()` also takes sequence-bearing objects directly — it
auto-detects them by S3 class, reads their stored sequences, and fits on
those, so the tree shares one symbol space with the source model. A
state-sequence `stslist` is one such object, fitted with no external
package:

```{r ecosystem}
context_tree(engagement, max_depth = 2L)     # an stslist of states
```

A fitted transition/network object that carries its underlying sequences
works the same way — pass it straight to `context_tree()`:

```{r ecosystem-net, eval = FALSE}
# transitiontrees also accepts a fitted transition/network object directly,
# auto-detected by its S3 class:
# context_tree(my_network_object, max_depth = 2L)
```

> When a fitted object carries its sequences, those are read and its node
> labels become the alphabet. A *pure* aggregated transition network
> (edges/weights only, no sequences) is rejected — transitiontrees fits on
> sequences, never on aggregated transitions.

# Session info

```{r session}
sessionInfo()
```
