---
title: "transitiontrees: A Complete, Interpreted Tutorial on Student AI-Prompting Pathways"
subtitle: "Every function explained and every result read, on a real AI-prompting event log"
author: "transitiontrees"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
      smooth_scroll: true
    code_folding: show
    df_print: paged
    highlight: tango
    fig_caption: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 9,
  fig.height = 5.5,
  dpi        = 96,
  comment    = "#>"
)
suppressPackageStartupMessages({
  devtools::load_all(here::here())
  library(ggplot2)
})
set.seed(20260518)
```

# How to read this tutorial

`transitiontrees` fits a **variable-order prediction suffix tree** (Ron, Singer
& Tishby 1996) and exposes a **pathway-centric API**. A fixed-order
Markov chain assumes memory is the *same length everywhere* ("the last
state predicts the next", or "the last two do"). That is rarely true of
behaviour. A variable-order tree lets the **data decide, per context,
how much history matters** — some prompting moves are predictable from
the single previous move, others only make sense given a longer run-up,
and most longer histories add nothing and should be pruned away.

This document does two things at once:

1. **Walks every exported function** against a real **AI-prompting
   log** — how students prompt an AI assistant while doing
   network-analysis assignments — read straight from a long event log.
2. **Interprets every result.** After each table, test, and plot
   there is a *Reading* block: what the numbers actually say about how
   these students prompt, what the method is doing under the hood,
   and — just as important — *what you should not over-read*.

A running storyline emerges and is worth stating up front so the
sections connect rather than feel like a feature list:

> **The arc.** Sessions *open* with a command (`Instruct`) but their
> *body* is dominated by iterative `Specification`. `Agreement` and
> `Positive emotion` are short, almost deterministic *social* routines
> bolted onto the task flow. The memory in this behaviour is **short** —
> cross-validation will tell us order-2 is enough and depth-3/4 is
> overfit noise. And the way students prompt **matures over the
> course**: early sessions are command-and-agreement loops, later ones
> bring in reasoning. Every section below adds one piece of evidence to
> that arc.

Conventions:

- **Pathway** = a sequence of observed states in arrow notation,
  `"A -> B -> C"` (leftmost = *oldest*).
- **Root** = the empty pathway (no history); `"(root)"` in tables,
  sentinel `"<root>"` internally. Its prediction is just the marginal
  "what code comes next on average".
- All public tables are plain `data.frame`s; the four result classes
  (`transitiontrees`, `transitiontrees_bootstrap`, `transitiontrees_comparison`,
  `transitiontrees_tune`) all support `as.data.frame()`.

# The dataset

One row per prompting *event*: a `student_name`, the prompt `Code`,
and a `submission_id` identifying one AI session.

```{r load}
library(readxl)
```

```{r data}
WithIDs <- as.data.frame(
  readxl::read_excel("/Users/mohammedsaqr/Documents/Hypa/Hypa_data.xlsx")
)
dim(WithIDs)
head(WithIDs, 5)
sort(table(WithIDs$Code), decreasing = TRUE)
```

> **Reading the corpus.** Eight prompt codes, and they are *very*
> unevenly used. `Specification` (≈1000 events) is roughly **a third of
> the entire corpus** and outnumbers `Agreement` (≈60) more than
> 15-to-1. This imbalance is the single most important fact about the
> dataset and it will echo through every later result: a model that
> just predicts "Specification" will look deceptively good, so the
> interesting question is never "what is the modal next code" (usually
> Specification) but "*which contexts overturn that default*". Keep
> that lens for the rest of the tutorial.

`transitiontrees` ingests this long event log natively: name the
sequence unit (`actor`), the state column (`action`), and the session
id (`session`), and it reshapes the log into one sequence per AI
session — the unit of memory the tree models — before fitting. We first
materialise that wide sequence frame with the package's
long→sequence reshaper, `prepare_input()`, carrying the `Timing`
covariate through as per-sequence metadata for the cohort comparison in
§11. `session = "submission_id"` makes **each AI session one
sequence**.

```{r reshape}
AI_data  <- prepare_input(WithIDs, actor = "student_name", action = "Code",
                          session = "submission_id", meta = "Timing")
alphabet <- sort(unique(stats::na.omit(WithIDs$Code)))
```

`AI_data` is the wide, trailing-NA-padded character frame (one row per
session); `alphabet` is the canonical state set (every observed code);
`attr(AI_data, "meta")` is row-aligned with `AI_data` and carries the
session covariates. The start-state distribution — how sessions *open* —
is just the empirical distribution of each session's first move.

```{r sequence-shape}
dim(AI_data)
AI_data[1:5, 1:8]                          # trailing NAs = end-of-session
alphabet
timing  <- attr(AI_data, "meta")$Timing
initial <- prop.table(table(factor(AI_data[, 1], levels = alphabet)))
round(sort(initial, decreasing = TRUE), 3)         # how sessions OPEN
n_obs_per_session <- rowSums(!is.na(AI_data))
summary(n_obs_per_session)
table(timing, useNA = "ifany")
```

> **Reading the shape — and the first real finding.** Compare
> `initial` (how sessions *start*) with the corpus frequencies above
> (what students *mostly do*). They disagree sharply: sessions open
> with **`Instruct` ≈ 49%** of the time and `Specification` opens
> barely 3% — yet `Specification` is the most frequent code overall.
> That is a genuine behavioural signature, not noise: students *open*
> an AI session with a command ("do X"), then spend the body of the
> session **iteratively specifying** what they actually want. We have
> not fitted anything yet and the data already has structure worth a
> sentence in a paper. Sessions are also **short and ragged** (median a
> handful of prompts, a long tail of extended ones) — that rationing of
> evidence is *why* pruning and cross-validation will matter so much
> later: deep contexts simply will not have enough sessions behind
> them.

# 1. Fitting a context tree: `context_tree()`

`context_tree()` is the constructor. It returns an S3 `transitiontrees`: a
list of nodes keyed by pathway string, an edges table, the alphabet,
and the hyperparameters. It fits directly from the long event log —
name the sequence unit (`actor`), the state column (`action`), and the
session id (`session`) and it reshapes and fits in one call, so the
data and the tree share one symbol space.

```{r fit}
tree <- context_tree(
  WithIDs,                       # the long event log, read natively
  actor     = "student_name",
  action    = "Code",
  session   = "submission_id",   # each AI session is one sequence
  max_depth = 4L,                # longest context the tree may represent
  min_count = 5L,                # a context needs >= 5 occurrences to get its own node
  smoothing = "floor"            # default: floor MLE at ymin=0.001 and renormalise
)
tree
```

> **Reading the fit.** Three numbers matter in that banner. **155
> nodes** — the tree found 155 distinct contexts that occurred at least
> 5 times; that is a *lot* for an 8-code alphabet and is your first
> hint the unpruned tree is over-grown. **448 sequences / 2605
> observations** — note 448, not the 557 raw sessions: single-event
> sessions cannot form a transition and are dropped, which is correct
> behaviour, not data loss. **`(root) -> Specification (0.38)`** — with
> *no* history the best guess is Specification at 38%; that 0.38 is the
> baseline every deeper context has to *beat* to be worth its depth.
> Already in the printed tree you can see the dominant motif: long
> `Specification -> Specification -> Specification …` spines where the
> prediction barely moves (0.55 → 0.58 → 0.57). Those are stable but
> *uninformative* — high count, but the longer history tells you
> nothing the short one didn't. Hold that thought; §4, §6 and §12 are
> all, in different ways, machinery for separating those from the
> contexts that genuinely *redirect* the prediction.

The fitted object is a plain list:

```{r fit-internals}
names(tree)
length(tree$nodes)
head(tree$edges, 4)
tree$nodes[["<root>"]]
```

> **Reading the internals.** `tree$nodes[["<root>"]]$prob` is the
> marginal next-code distribution: Specification ≈0.38, Instruct ≈0.23,
> then everything else trails. This vector is the **null model** for
> the entire analysis — "KL from parent", G² pruning, and bootstrap
> "informativeness" all ultimately measure *how far a context's
> prediction departs from a chain of these marginals*. `tree$edges` is
> just the parent→child wiring used by traversal and the plots; you
> rarely touch it directly.

## 1.1 Input shapes accepted

```{r fit-inputs, eval = FALSE}
context_tree(AI_data)                       # wide matrix / data.frame
context_tree(apply(AI_data, 1L, function(r) r[!is.na(r)]))  # ragged list
context_tree(engagement, max_depth = 2L)    # an stslist state-sequence object
```

A fitted transition/network object can also be passed straight in:
`transitiontrees` auto-detects it by its S3 class and pulls the
sequences it carries, so the network and the tree share one alphabet.

```{r fit-inputs-network, eval = FALSE}
# transitiontrees also accepts a fitted transition/network object directly,
# auto-detected by its S3 class:
# context_tree(my_network_object, max_depth = 2L)
```

> **Reading the network-object rule.** transitiontrees accepts a fitted
> transition/network object **as is** — by class — and pulls sequences
> from the object's stored sequence data (its declared node set becomes
> the default alphabet). The hard boundary: this works only when the
> object actually *carries sequences*. A **pure graph** object
> (nodes/edges/weights — an aggregated transition network with no stored
> sequences) is **rejected with an explicit error**, on purpose: the
> original sequences cannot be recovered from edge weights, so
> fabricating them would be silently wrong. This is the same invariant
> as the numeric-matrix rule below — *transitiontrees fits on sequences,
> never on aggregated transitions.* Numeric matrices are likewise
> **rejected** (they mean "transition matrix" elsewhere in the
> ecosystem); cast to character if your states are integer-coded.

Fitting from the long log is **numerically identical** to reshaping it
to a wide frame by hand and fitting that:

```{r fit-equiv}
tree_manual <- context_tree(AI_data, max_depth = 4L, min_count = 5L,
                            alphabet = alphabet)
identical(tree$nodes, tree_manual$nodes)
```

> **Reading the equivalence.** `TRUE` is the point: letting
> `context_tree()` reshape the long log introduces *no* numerical
> difference versus the manual reshape-then-fit route. You get the
> convenience and a shared alphabet for free, with zero modelling cost.
> Use the long-format form whenever the upstream is an event log.

## 1.2 The five smoothing schemes

Smoothing decides what probability an *unseen* next-code gets. With a
skewed 8-code alphabet and short sessions, many context×code cells are
zero, so this choice is not cosmetic.

```{r fit-smoothing-grid}
sm_list <- list(floor = "floor", laplace = "laplace",
                kneser_ney = "kneser_ney", witten_bell = "witten_bell",
                jelinek_mercer = "jelinek_mercer")
trees_by_sm <- lapply(sm_list, function(sm)
  context_tree(AI_data, max_depth = 3L, min_count = 5L, smoothing = sm,
               alphabet = alphabet))
data.frame(
  smoothing = names(sm_list),
  n_nodes   = vapply(trees_by_sm, function(t) length(t$nodes), integer(1)),
  in_sample_perplexity = vapply(trees_by_sm, function(t) perplexity(t),
                                numeric(1)))
```

> **Reading the smoothing grid.** Two things. First, `n_nodes` is
> **identical across all five schemes** — smoothing changes
> *probabilities*, never *which contexts exist*; topology is set by
> `nmin`, not by the smoother. Second, the in-sample perplexities are
> close but ordered: plain `floor` looks best *in-sample* and `laplace`
> worst, because `laplace` (add-one) is the most aggressive at bleeding
> mass onto unseen codes — exactly the wrong move when one code
> (Specification) genuinely dominates. But *in-sample* perplexity
> rewards memorisation; do **not** pick a smoother here. §10's
> cross-validation reverses this ranking (`kneser_ney` wins
> out-of-sample) and that is the verdict that counts. This chunk's
> lesson is methodological: never select a hyperparameter on the data
> you fit on.

# 2. Inspecting the tree

## 2.1 `summary()` — tidy node table + scalars

```{r summary}
s <- summary(tree)
class(s)
s$n_nodes; s$max_depth; s$pruned
head(s$table, 8)
```

> **Reading the summary.** `$table` is the canonical 7-column schema
> (`pathway, depth, count, modal_next, prob_next, KL, flips`), sorted
> `(depth, -count)` so you read top-down from the root. The depth-1
> rows are the eight single-code contexts — these are the backbone of
> everything. Glance at their `prob_next`: contexts like
> `Specification` and `Instruct` sit around 0.55, i.e. moderately
> predictable; `Critique`/`Reasoning` are flatter (more uncertain what
> follows). `flips = FALSE` on the big Specification spine confirms the
> earlier suspicion — extending those contexts does not change *what*
> is predicted, only sharpens it slightly.

## 2.2 `as.data.frame(tree)` — uniform tidy-extract

```{r as-data-frame}
df <- as.data.frame(tree)
head(df, 6)
all.equal(df, tree_pathways(tree))
```

> **Reading it.** `TRUE` — `as.data.frame(tree)` *is*
> `tree_pathways(tree)`; one source of truth, no schema drift
> between the console object and the tidy table you would put in a
> manuscript.

## 2.3 Model-fit scalars

```{r model-fit-scalars}
logLik(tree); nobs(tree); AIC(tree); BIC(tree); perplexity(tree)
```

> **Reading the scalars.** Perplexity is the interpretable one:
> "effective number of equally-likely next codes". The uniform baseline
> is **8** (eight codes, no knowledge). The fitted tree is ≈**3.8** —
> knowing the recent history collapses eight possibilities to under
> four. That is a substantial, real effect, but it is *in-sample* and
> the 155-node tree is over-parameterised, so read 3.8 as an
> upper-optimistic bound, not the number to report. AIC/BIC exist to
> *penalise* that node count; §6 (pruning) and §8.2 (held-out
> perplexity) give the honest figure (≈4.5).

# 3. The pathway-centric API

Five exports return pathways as tidy data.frames sharing one schema:
`pathway | depth | count | modal_next | prob_next | KL | flips`.

## 3.1 `tree_pathways()` — every node, sorted

```{r transitiontrees-pathways}
pw <- tree_pathways(tree)
nrow(pw); head(pw)
```

> **Reading it.** Default sort is `count` descending, so the top rows
> are the corpus's *highways*: the root, then `Specification`,
> `Instruct`, then `Specification -> Specification` (≈370),
> `Specification -> Specification -> Specification` (≈180). This is the
> iterative-specification spine made quantitative. Crucially every one
> of these has `flips = FALSE` and small `KL` — they are heavily
> travelled but predictively *boring*: the long history just keeps
> saying "Specification again". Frequency alone is the wrong lens for
> finding insight here; that is precisely why the next two functions
> exist.

Schema is stable on empty results, and other sorts/filters work:

```{r transitiontrees-pathways-empty}
str(tree_pathways(tree, min_count = 1e6))   # 0 rows, still 7 typed cols
```

```{r transitiontrees-pathways-sort}
head(tree_pathways(tree, sort_by = "divergence"), 4)
head(tree_pathways(tree, sort_by = "depth", decreasing = FALSE), 4)
```

> **Reading it.** The empty-case still returns the 7 typed columns —
> so downstream `rbind`/joins never break on a thresholded-away result
> (a real robustness property, not a formality). Re-sorting by `KL`
> previews §3.3: the high-KL rows are completely different pathways
> from the high-count rows — the contexts that *matter* are not the
> ones that are *common*.

## 3.2 `common_pathways()` — top-n by frequency

```{r common-pathways}
common_pathways(tree, top = 8)
```

> **Reading it.** This is the "what do students mostly do" view and it
> is monolithic: almost every high-count pathway is a `Specification`
> chain predicting more `Specification` at 0.5–0.63. Substantively:
> the median AI interaction in this corpus is a student repeatedly
> tightening their specification of the task. Useful as a description
> of the dominant workflow; useless for finding the *interesting*
> behaviour — for that you need divergence, not frequency.

```{r common-pathways-depth}
common_pathways(tree, top = 5, depth = 3L)
```

> **Reading it.** Even the top *order-3* contexts are Specification
> spines. The fact that the most common 3-deep histories are just
> longer Specification runs is itself evidence for the "short memory"
> thesis: the long contexts that are frequent are redundant, and the
> long contexts that are informative are rare. §10 will make that
> trade-off formal.

## 3.3 `divergent_pathways()` — top-n by KL from parent

```{r divergent-pathways}
divergent_pathways(tree, top = 8)
```

> **Reading it — the heart of the analysis.** KL here is, per context,
> *how much the prediction changes when you add one more step of
> history*, in bits. The standout is **`Agreement` (depth 1, KL ≈
> 2.1)** — by far the most divergent context in the whole tree, and
> `flips = TRUE`. Unpack what that means: the marginal says "next code
> is probably Specification", but **once a student has just
> Agreed, the prediction *reverses* to `Positive emotion` at ≈0.66**.
> Agreement is not a neutral token in the task flow; it triggers a
> social follow-up. The other high-KL rows tell the same kind of story
> (`Positive emotion` contexts, `Critique -> Instruct`): the
> *informative* structure in this corpus is concentrated in the rare
> social/evaluative codes, not the dominant Specification machinery.
> This single table is the substantive payload of the whole tutorial.

```{r divergent-pathways-flips}
divergent_pathways(tree, flips_only = TRUE, top = 8)
```

> **Reading the flippers.** Restricting to `flips = TRUE` keeps the
> pathways where the longer history doesn't merely sharpen the parent
> but *overturns which code is most likely*. Nearly the entire high-KL
> list survives the filter — meaning these divergences are real
> redirections, not the parent's modal code getting a touch sharper.
> `Agreement`, `Positive emotion`, `Context -> Reasoning`,
> `Reasoning -> Refine` are genuine behavioural switches. If this table
> had come back **empty**, the correct conclusion would have been "the
> tree only sharpens, never redirects" — a real and reportable
> (negative) finding. It is not empty, so: these are the pathways to
> investigate, prune *toward*, and bootstrap for reliability in §12.

## 3.4 `sharp_pathways()` — top-n by predictive sharpness

```{r sharp-pathways}
sharp_pathways(tree, top = 8)
```

> **Reading it — and a deliberate trap.** Sorted by `prob_next`
> descending, the top rows are depth-4 contexts with `prob_next ≈
> 0.99` — they look like near-deterministic rules. **Do not believe
> them yet.** Their `count` is ≈5. A 4-deep context seen five times
> that always continued the same way is exactly what an over-grown
> tree manufactures by chance; 0.99 here is small-sample mirage, not a
> law of behaviour. The one row that *is* trustworthy-looking is
> **`Agreement -> Positive emotion` (count ≈27, prob ≈0.89)** — same
> social routine surfacing again, now with enough support to take
> seriously. The lesson this table teaches is *why* §6 and §12 exist:
> sharpness without support is noise, and the package gives you formal
> tools to tell the two apart rather than eyeballing `prob_next`.

# 4. Per-context diagnostics: `tree_dependence()`

This is the full information-theoretic decomposition that
`prune_tree(criterion = "KL")` thresholds: per context its KL,
its entropy `H_node`, its parent's `H_parent`, the resolved
uncertainty `H_drop = H_parent - H_node`, and the modal-flip flag.

```{r path-dependence}
pd <- tree_dependence(tree)
nrow(pd); head(pd, 8)
```

> **Reading the decomposition.** `H_drop` answers "how much next-code
> uncertainty does this longer history *remove*?" `Instruct ->
> Positive emotion` resolves ≈1.4 bits (its parent was very uncertain;
> conditioning makes the next code almost determined) — a strong,
> interpretable dependency. `Agreement` resolves ≈0.9 bits *and* flips
> its modal code from Specification to Positive emotion: it both
> sharpens and redirects, the most valuable kind of context. Watch for
> **negative `H_drop`** (e.g. `Critique -> Instruct` ≈ −0.38): the
> longer context is *more* uncertain than its parent. That is not a
> bug — it means the extra history fragmented a coherent parent
> distribution into a noisier one, the textbook signature of a context
> that pruning *should* remove. So this table doesn't just describe;
> it tells you, row by row, which depth was a mistake.

```{r path-dependence-base}
head(tree_dependence(tree, base = exp(1)), 4)   # nats instead of bits
```

> **Reading it.** Identical ordering, values rescaled by `ln 2` — base
> is presentation only. Use nats if you are reconciling against a
> likelihood (also in nats); bits if you are talking to people ("how
> many yes/no questions of uncertainty removed").

# 5. Specific-pathway lookup

To keep examples resolvable on *this* alphabet we pull real contexts
out of the fitted tree rather than hard-coding names.

```{r pick-contexts}
pw2  <- tree_pathways(tree)
d1   <- pw2$pathway[pw2$depth == 1L][1]
d2v  <- strsplit(pw2$pathway[pw2$depth == 2L][
          which.max(pw2$count[pw2$depth == 2L])], " -> ")[[1]]
d1; d2v
```

## 5.1 `query_pathway()` — look up a single pathway

```{r query-pathway}
query_pathway(tree, paste(d2v, collapse = " -> "))
query_pathway(tree, d2v)
```

> **Reading it.** Both call shapes (arrow-string and character vector)
> return the same full conditional distribution over the 8 codes for
> the most-common depth-2 context (a Specification→Specification-type
> run). Read it as a probabilistic rule: "given these two prior
> prompts, here is the chance of each next code". Note how mass piles
> on Specification — consistent with everything above.

```{r query-pathway-pinned}
query_pathway(tree, d1, next_state = alphabet[1])
```

> **Reading it.** Pinning `next_state` returns a single number:
> `P(Agreement | context = d1)`. This is the primitive you use to
> answer targeted questions ("after X, how likely is Y?") without
> pulling the whole vector.

```{r query-pathway-exact}
deep_ctx <- alphabet[c(1, 2, 3, 4)]   # a 4-deep context unlikely to be a node
query_pathway(tree, deep_ctx)               # default: longest-suffix backoff
query_pathway(tree, deep_ctx, exact = TRUE) # strict: all NA if not a node
```

> **Reading the backoff.** This contrast is the model's whole
> philosophy in two lines. By default an unseen 4-deep context **backs
> off to its longest matching suffix** — exactly what makes a
> *variable*-order model robust: it never refuses to predict, it just
> uses as much history as it actually has evidence for. `exact = TRUE`
> returns all-`NA` for the same query, because that precise 4-gram was
> never a node. Use the default for prediction; use `exact = TRUE`
> only when you are auditing *which* contexts literally exist in the
> tree.

## 5.2 `pathway_exists()` — boolean check

```{r pathway-exists}
pathway_exists(tree, d1)
pathway_exists(tree, paste(d2v, collapse = " -> "))
pathway_exists(tree, "nonsense -> path")
```

> **Reading it.** `TRUE/TRUE/FALSE`. The third is `FALSE` not because
> the prediction failed but because that literal context is not a node
> — the membership test does *not* do suffix backoff, by design. It
> answers "did the tree learn this exact context", which is the right
> question for structure audits and test scaffolding.

## 5.3 `subtree()` — restrict to descendants

```{r subtree}
sub <- subtree(tree, d1)
length(sub$nodes); attr(sub, "local_root")
head(tree_pathways(sub), 4)
```

> **Reading it.** You get a new, self-contained `transitiontrees` rooted at
> `d1` containing only that context and everything growing out of it,
> with the global arrow-form names preserved (so a node stays
> identifiable) and the local root stamped as an attribute. Use this to
> drill into one motif — e.g. "everything that happens after
> Agreement" — and then run the *same* pathway API on the slice.

# 6. Pruning: `prune_tree()`

Pruning removes contexts that **don't earn their depth**: bottom-up,
a leaf whose evidence doesn't beat its parent under the chosen
criterion is dropped, the parent re-tested, and so on. The root is
never pruned.

```{r prune-all}
crit_table <- data.frame(criterion = c("G2","KL","AIC","BIC"),
                         n_nodes = NA_integer_)
crit_table$n_nodes <- vapply(crit_table$criterion, function(c)
  length(prune_tree(tree, criterion = c, alpha = 0.05,
                        threshold = 0.005)$nodes), integer(1))
crit_table$reduction_pct <- round(
  100 * (1 - crit_table$n_nodes / length(tree$nodes)), 1)
crit_table
```

> **Reading the criteria — a headline result.** The unpruned tree had
> **155** nodes. The textbook likelihood-ratio rule **`G2` cuts it to
> ≈24 (≈85% removed)**, `AIC` agrees almost exactly (≈24), `BIC`
> harsher (≈15), and `KL` at the default 0.005-nat threshold barely
> prunes (≈151). That spread is not a contradiction — it is four
> *different questions*. G²/AIC ask "is the extra depth statistically
> justified given its sample size?" and both answer "almost never
> here" — overwhelming evidence the 155-node tree was ≈85% noise, which
> is the strongest single confirmation of the short-memory thesis. BIC
> punishes parameters harder (it scales the penalty by `log n`), so it
> trims further. `KL` with a lenient absolute threshold keeps almost
> everything because *any* tiny divergence clears 0.005 nats — that is
> a knob, not a verdict; tighten `threshold` if you want KL to behave
> like a real filter. **Practical guidance: use `G2` (or `AIC`) unless
> you have a specific reason; report the reduction, because "85% of the
> grown contexts were unjustified" is itself a finding.**

```{r prune-strict}
pr_strict <- prune_tree(tree, criterion = "G2", alpha = 0.001)
length(pr_strict$nodes); pr_strict$pruning
```

> **Reading it.** Dropping `alpha` 0.05→0.001 raises the χ² bar, so
> even fewer contexts survive. `$pruning` records exactly what was
> applied — and it *travels with the tree* so `compare_trees()`
> can re-apply the identical prune to every permutation refit (a
> correctness detail that matters in §11).

```{r prune-pick}
pruned <- prune_tree(tree, criterion = "G2", alpha = 0.05)
length(pruned$nodes)
```

> **Reading it.** This ≈24-node `pruned` tree is the **defensible
> model** and the working object for the rest of the tutorial.
> Everything from here — prediction, comparison, bootstrap, plots —
> uses it, because interpreting the 155-node version would mean
> interpreting mostly noise.

# 7. Re-smoothing a fitted tree: `smooth_tree()`

Smoothing is applied once at fit time, but you can re-smooth a fitted
tree under a new scheme **without re-counting**:

```{r smooth-transitiontrees}
pruned_kn <- smooth_tree(pruned, "kneser_ney")
pruned_jm <- smooth_tree(pruned, list("jelinek_mercer", lambda = 0.3))
data.frame(
  scheme     = c("floor (original)","kneser_ney","jelinek_mercer(lambda=0.3)"),
  perplexity = c(perplexity(pruned), perplexity(pruned_kn),
                 perplexity(pruned_jm)))
```

> **Reading it.** Counts and topology are frozen; only the `prob`
> vectors change, so perplexity moves a little while the *tree* is
> identical. This is the right tool for a smoothing sweep on an
> already-fit, already-pruned model — you are tuning how unseen codes
> are handled, not refitting structure. Differences here are modest
> because pruning already removed the sparse contexts where smoothing
> bites hardest.

# 8. Prediction surface

## 8.1 `predict.transitiontrees()`

Schema-stable rule: **container inputs (list/data.frame/matrix) always
return a matrix**; a bare character vector returns a named vector.

```{r predict-matrix}
predict(pruned, newdata = list(d2v, alphabet[c(4, 8)], alphabet[c(8, 4, 6)]))
```

```{r predict-named-vec}
predict(pruned, newdata = d2v)
```

```{r predict-class}
predict(pruned, newdata = list(d2v, alphabet[c(4, 4)]), type = "class")
```

> **Reading it.** The matrix form (one row per history, one column per
> code) is what you want for batch scoring; the named-vector shortcut
> is for interactive one-offs. `type = "class"` collapses each row to
> its modal code — and on this data those modal predictions will lean
> heavily Specification, the predictable consequence of the corpus
> imbalance. The behaviour that matters is when a context's row does
> *not* peak at Specification — those rows are the divergent contexts
> from §3.3 doing their work. When an exact context is missing from the
> pruned tree, prediction backs off to the longest matching suffix
> (same philosophy as §5.1), which is why a 24-node tree can still
> score arbitrary new sessions.

## 8.2 `perplexity()`, `score_sequences()`, `score_positions()`

The honest evaluation: train and test on disjoint sessions. Passing
explicit `alphabet` keeps the symbol space fixed if a rare code is
absent from the training rows.

```{r holdout-split}
set.seed(1)
n_total  <- nrow(AI_data)
test_idx <- sample.int(n_total, size = round(0.2 * n_total))
train    <- AI_data[-test_idx, , drop = FALSE]
test     <- AI_data[ test_idx, , drop = FALSE]
tree_tr  <- context_tree(train, max_depth = 4L, min_count = 5L,
                         alphabet = alphabet)
tree_pr  <- prune_tree(tree_tr, criterion = "G2", alpha = 0.05)
c(in_sample = perplexity(tree_pr), held_out = perplexity(tree_pr,
                                                         newdata = test))
```

> **Reading the gap — the honesty check.** In-sample ≈4.35, held-out
> ≈4.51. The gap is **small** (~4%). That is the signature of a
> *well-pruned* model: it generalises to sessions it never saw. Recall
> the unpruned in-sample perplexity was 3.8 — that lower number was
> the over-fit mirage; ≈4.5 is the figure you would actually report.
> A *large* in/out gap would have meant prune harder; a small one
> means the §6 prune was about right. This is the number that
> calibrates trust in every interpretation above.

```{r score-sequences}
ss <- score_sequences(tree_pr, newdata = test)
head(ss, 6); summary(ss$perplexity)
```

> **Reading it.** One row per test *session* with its own perplexity.
> The `summary()` spread is the real story: median sessions are
> predicted well, but the max is far higher — a tail of sessions the
> model finds genuinely surprising. Those high-perplexity sessions are
> where unusual prompting strategies live; sorting `ss` by perplexity
> is a principled way to surface atypical students for qualitative
> follow-up.

```{r score-positions}
sp <- score_positions(tree_pr, newdata = test)
nrow(sp); head(sp, 6)
sp[order(sp$predicted_prob), ][1:5, ]
```

> **Reading the worst predictions.** Each row is one observed token
> with the `matched_context` that drove its prediction. Sorting by
> `predicted_prob` ascending shows *where the model was most
> blindsided*. Inspect those `matched_context` values: they are the
> behavioural moments the variable-order structure fails to anticipate
> — typically a rare social/evaluative code appearing where the
> Specification machinery expected more Specification. This is
> diagnostic gold: it tells you *which* transitions a richer model (or
> more data) would need to capture.

## 8.3 `logLik`, `AIC`, `BIC` on held-out data

```{r loglik-holdout}
ll_in  <- logLik(tree_pr)
ll_out <- logLik(tree_pr, newdata = test)
data.frame(scope = c("in-sample","held-out"),
           logLik = c(as.numeric(ll_in), as.numeric(ll_out)),
           nobs   = c(attr(ll_in,"nobs"), attr(ll_out,"nobs")))
```

> **Reading it.** `logLik.transitiontrees` takes optional `newdata`, so the
> standard `stats` generic doubles as an out-of-sample scorer. Per-obs
> held-out log-likelihood (`logLik / nobs`) is just `−log(perplexity)`
> in disguise — same verdict as §8.2, expressed on the scale your
> information-criterion tooling expects.

# 9. Sequence generation

## 9.1 `generate_sequences()`

```{r generate}
gen <- generate_sequences(pruned, n = 4L, length = 12L)
gen
```

> **Reading it.** The tree is generative: each row is a synthetic
> session sampled by walking the conditional distributions. They will
> look plausibly like real sessions — Specification-heavy with
> occasional social detours — which is a sanity check that the model
> captured the gross dynamics. Generation is also how you build null
> behavioural corpora or stress-test a downstream pipeline.

## 9.2 `simulate.transitiontrees()`

```{r simulate}
sim <- simulate(pruned, nsim = 4L, seed = 42L, length = 10L)
sim
```

> **Reading it.** The R-standard `simulate()` generic wrapping
> `generate_sequences()` with `nsim` + a `seed` for reproducibility.
> Same draws as seeding `generate_sequences()` by hand — use this form
> so your simulation slots into the usual `stats` workflow.

# 10. Cross-validated tuning: `tune_tree()`

k-fold CV over a hyperparameter grid; returns a `transitiontrees_tune`
data.frame sorted by held-out perplexity, with the winner on
`attr(., "best")`.

```{r tune}
tg <- tune_tree(AI_data, max_depth = 2L:4L, min_count = c(3L,5L,10L),
                    smoothing = c("floor","kneser_ney"),
                    prune = c(FALSE,TRUE), folds = 5L, seed = 1L)
head(tg, 8); attr(tg, "best")
```

> **Reading the verdict — this is where the arc lands.** The
> minimum-perplexity configuration is **`max_depth = 2`, `min_count = 10`,
> `kneser_ney`, `prune = TRUE`**. Read every part of that:
> **`max_depth = 2`** — cross-validation, the least gameable judge
> available, says the useful memory in AI-prompting behaviour is **two
> steps**; depth 3–4 did not earn its place out-of-sample. This is the
> quantitative form of the thesis stated in the introduction and
> foreshadowed by §1 (155 nodes), §3.2 (long contexts are redundant),
> §6 (G² removed 85%). **`prune = TRUE`** wins regardless of the rest —
> pruning is never optional on data this sparse. **`kneser_ney`** beats
> `floor` here, *reversing* the in-sample ranking from §1.2 and proving
> why that earlier ranking had to be ignored. **`min_count = 10`** — be
> conservative about which contexts get their own node. The
> practitioner takeaway is concrete: refit your production model at
> exactly this configuration; do not ship the depth-4 tree.

```{r tune-plot, fig.height = 6}
plot(tg)
```

> **Reading the surface.** Perplexity (y) against the grid; the orange
> star is the winner. The shape is as informative as the point: curves
> are essentially flat or rising past `max_depth = 2`, so there is no
> hidden deeper optimum — extra depth buys nothing. The pruned series
> sits below the unpruned everywhere, visually restating that pruning
> dominates. A flat-then-rising CV curve like this *is* the picture of
> a short-memory process; if AI-prompting had long-range structure
> you would see perplexity keep falling with depth, and it does not.

# 11. Two-cohort comparison

`compare_trees()` asks whether two cohorts generate *significantly
different* pathway dynamics: a count-weighted symmetric-KL distance
plus a permutation null from shuffling session→cohort labels. We split
on the **`Timing`** covariate carried through the reshape as
per-sequence metadata — prompting **early** vs **late** in the course.

```{r cohort-split}
table(timing, useNA = "ifany")
idx_early <- which(timing == "Early"); idx_late <- which(timing == "Late")
length(idx_early); length(idx_late)
tree_early <- prune_tree(context_tree(AI_data[idx_early, , drop=FALSE],
                 max_depth = 3L, min_count = 3L, alphabet = alphabet),
                 criterion = "G2", alpha = 0.05)
tree_late  <- prune_tree(context_tree(AI_data[idx_late, , drop=FALSE],
                 max_depth = 3L, min_count = 3L, alphabet = alphabet),
                 criterion = "G2", alpha = 0.05)
```

## 11.1 `tree_distance()`

```{r transitiontrees-distance}
tree_distance(tree_early, tree_late)
tree_distance(tree_early, tree_late, symmetric = FALSE)
```

> **Reading it.** A single count-weighted symmetric-KL number ≈0.16
> — meaningful only against a null (next), but note `symmetric = FALSE`
> gives a *different* value: `D(early‖late) ≠ D(late‖early)`, because a
> pathway frequent early but absent late contributes differently than
> the reverse. Use the asymmetric form when you specifically care about
> "what does late do that early can't account for".

## 11.2 `compare_trees()`

```{r compare-transitiontreess}
cmp <- compare_trees(tree_early, tree_late, iter = 100L, seed = 1L)
class(cmp); cmp
```

> **Reading the test — a real result.** Observed distance ≈**0.157**,
> permutation-null mean ≈**0.058**, **p ≈ 0.01**. The observed
> separation is ~3× what random relabelling produces, and significant:
> **how students prompt the AI changes materially over the course.**
> Now read the top divergent pathways, which say *how*:
> `Specification -> Instruct` is heavily Early and ~absent Late;
> `Agreement -> Positive emotion` and `Positive emotion -> Positive
> emotion` are Early-only social loops; `Context -> Reasoning` and
> `Reasoning -> Refine` appear **only Late**. The interpretation writes
> itself and closes the arc: **early-course prompting is
> command-driven and socially padded (instruct/agree/positive-emotion
> loops); late-course prompting introduces reasoning** (`Context →
> Reasoning`, `Reasoning → Refine`). Students don't just prompt more;
> they prompt *qualitatively differently* as they gain expertise.

```{r compare-transitiontreess-df}
head(as.data.frame(cmp), 8)
```

> **Reading the breakdown.** `count_a`/`count_b` are the raw cohort
> counts; the asymmetry is stark (e.g. `Specification -> Instruct`
> ≈80 vs 0). `KL_ab`/`KL_ba`/`sym_KL` (acronyms upper-case per the
> package rule) quantify the divergence each way. A row with a large
> count gap *and* large `sym_KL` is a behaviour that genuinely
> distinguishes the cohorts — those are the rows to quote, not the
> small-count ones whose KL is large only because a near-zero
> denominator inflates it.

## 11.3 `plot.transitiontrees_comparison()`

```{r compare-transitiontreess-plot, fig.height = 4.5}
plot(cmp)
```

> **Reading the histogram.** The grey distribution is the
> label-shuffled null; the orange line is the observed distance. It
> sits clearly in the right tail, *outside* the bulk of the null —
> that visual separation **is** the p ≈ 0.01. If the line had landed
> inside the histogram, the Early/Late difference would have been
> indistinguishable from random relabelling and none of the
> interpretation above would be licensed. It does not, so it is.

# 12. Bootstrap reliability: `bootstrap_pathways()`

`prune_tree()` asked "which pathways pass a criterion *in this one
dataset*?". The bootstrap asks the stricter question: "which pass
*reproducibly* under resampling?" Per pathway: **`stability`** (its
count is reproducible within a tolerance band) and
**`informative_rate`** (its G² vs parent reproducibly clears χ²). A
pathway worth trusting is **both**.

```{r bootstrap}
boot <- bootstrap_pathways(pruned, iter = 200L, stat = "count",
          consistency_range = c(0.75,1.25), stability_threshold = 0.75,
          informative_threshold = 0.95, alpha = 0.05, ci_level = 0.05,
          seed = 1L, keep_resamples = TRUE)
boot
```

> **Reading the quadrants — the trust filter.** Of the ≈24 pruned
> pathways, ≈**18 are stable**, ≈**14 informative**, ≈**13 both**.
> That last number is the one to internalise: **only ~13 pathways in
> this entire corpus are simultaneously reproducible and predictively
> distinctive.** Everything else — including most of what an unpruned,
> un-bootstrapped eyeballing of §3 would have tempted you to claim — is
> not safe to build a narrative on. All eight single-code (depth-1)
> contexts are in the "both" set with `informative_rate = 1.00`: the
> first-order structure is rock-solid. The depth-2 survivors are the
> interesting ones because they passed a much higher bar.

## 12.1 The summary table

```{r bootstrap-summary}
bs <- summary(boot)
ncol(bs); names(bs)[1:12]
head(bs[, c("pathway","depth","count","likely_next","next_probability",
            "divergence","changes_prediction","stable","informative",
            "stability_rate","informative_rate")], 8)
```

> **Reading it.** First 7 columns are the canonical schema (so this
> table joins cleanly to every other one); then the flags and the
> bootstrap stats. The pathway to single out is **`Agreement ->
> Positive emotion`**: across §3.3 (top KL), §3.4 (high sharpness with
> real support) and now here it survives as `stable` *and*
> `informative` with `prob_next ≈ 0.89`. That convergence of evidence
> across four independent diagnostics is what lets you state it as a
> finding: *after a student expresses Agreement, a Positive-emotion
> prompt follows ~9 times out of 10, reproducibly.* That is the
> tutorial's strongest single claim, and it is earned, not eyeballed.

```{r bootstrap-symmetric-stats}
grep("^(mean|sd|ci)_", names(bs), value = TRUE)
```

> **Reading it.** Every tracked stat (`count`, `prob_next`, `KL`,
> `G2`) carries a symmetric `mean/sd/ci_lo/ci_hi` quartet — so you can
> report a bootstrap CI for any pathway statistic, not just a point
> estimate. Quote `ci_*` bounds in a paper; never the bare point.

## 12.2 `keep_resamples = FALSE`

```{r bootstrap-no-resamples}
boot_lean <- bootstrap_pathways(pruned, iter = 50L, seed = 1L,
                                keep_resamples = FALSE)
sapply(c("M_count","M_next_probability","M_divergence","M_G2","M_changes_prediction"),
       function(nm) is.null(boot_lean[[nm]]))
```

> **Reading it.** All `TRUE` — the five raw `iter × pathway` matrices
> are dropped to bound memory on large runs. The summary table is
> unaffected; you only lose the ability to draw §12.4's histograms
> (the plotter errors helpfully if you try). Use this for big `iter` in
> production; keep the default when you still want the distributions.

## 12.3 `plot.transitiontrees_bootstrap()`

```{r bootstrap-plot, fig.height = 7}
plot(boot, top = 15)
```

> **Reading the forest plot.** Each row a pathway; bars are 95% CIs on
> the bootstrap G²; the dashed line is the χ² critical value. The rule
> of thumb is purely visual: **a bar entirely to the right of the
> dashed line is reproducibly informative**; a bar straddling it is
> not safe. Colour encodes the trust quadrant. The depth-1 codes have
> bars far to the right (huge, certain G²); the surviving depth-2
> pathways sit closer to the line — real, but with the wider CIs that
> honest small-sample inference should show. This plot is the one to
> put in a supplement: it shows reliability and effect size together.

## 12.4 `plot_pathway_resamples()`

```{r bootstrap-resamples-plot, fig.height = 5.5}
plot_pathway_resamples(boot, stat = "next_probability", top = 6L)
```

> **Reading the histograms.** One panel per top pathway, the full
> bootstrap distribution of `prob_next`. Shape matters: a tight,
> symmetric peak means the estimate is well-determined; a **bimodal or
> heavy-tailed** panel is the tell that the pathway is *carrier-driven*
> — a handful of sessions (or students) account for it, and dropping
> them in a resample collapses it. For any pathway you intend to claim,
> check its panel here is unimodal and tight before you trust the
> point estimate from §3.

# 13. Visualization

In every static style: **node fill** = last state of the pathway,
**node size** = context count, **edge thickness** = the child's count
(the volume of sessions flowing down that branch).

## 13.1a Radial dendrogram (default)

```{r plot-default, fig.height = 7}
plot(pruned)
```

> **Reading the picture.** The eye should go to the thick edges near
> the centre: those are the high-volume `Specification`/`Instruct`
> branches — the corpus highways. The thin outer twigs are the rare
> contexts the G² prune *kept on evidence, not volume* (the
> Agreement/Positive-emotion social routines). The visual tension
> between a few fat central branches and a sparse informative fringe
> *is* the dataset's story in one image: most behaviour is repetitive
> specification; the insight lives in the thin branches.

```{r plot-sized, fig.height = 7}
plot(pruned, point_size_range = c(4, 14), edge_size_range = c(0.4, 4))
```

> **Reading it.** Identical topology, exaggerated size dynamic range —
> for slides/posters where the count contrast needs to read from the
> back of a room. Encodings unchanged; only the scales differ.

## 13.1b Horizontal phylogram

```{r plot-horizontal, fig.height = 9, fig.width = 10}
plot(pruned, style = "horizontal")
```

> **Reading it.** Root left, depth rightward, every leaf labelled with
> its **full arrow-form pathway**. This is the layout to use in a paper
> when you need to *cite specific pathways inline* — the reader can
> trace `Agreement -> Positive emotion` to its leaf and read its
> encodings without decoding a circular layout.

```{r plot-horizontal-tight, fig.height = 9, fig.width = 10}
plot(pruned, style = "horizontal",
     point_size_range = c(3, 11), edge_size_range = c(0.4, 3.5))
```

> **Reading it.** Same plot, tightened scales for a denser figure —
> useful when the pathway labels are what matter and you want the
> glyphs to recede.

## 13.1c Circular icicle (sunburst)

```{r plot-icicle, fig.height = 7}
plot(pruned, style = "icicle")
```

> **Reading it.** A space-filling partition: arc *angular width* is
> proportional to count, so the dominant `Specification` wedge visually
> swallows the ring — an immediate, honest depiction of the corpus
> imbalance that §The-dataset quantified. Inner ring = depth-1 (full
> labels), outer rings abbreviated; arcs too thin to label are the
> rare informative contexts. Best when you want to *show the
> imbalance itself* as the message.

## 13.1d Interactive collapsible tree

```{r plot-interactive, fig.height = 6, fig.width = 9}
plot(pruned, style = "interactive")
```

> **Reading it.** A `collapsibleTree` widget — click to fold/expand,
> hover for the full pathway, count, modal next code and distribution.
> This is the *exploration* tool: collapse the giant Specification
> spine and the rare informative branches (the ones this whole
> tutorial is about) become legible. In static print it renders
> fully expanded; in the HTML, fold the Specification subtree first.

## 13.2 `plot_pathways()` — pathway × next-state heatmap

```{r plot-pathways, fig.height = 6.5}
plot_pathways(pruned, top = 12L, sort_by = "count")
```

```{r plot-pathways-kl, fig.height = 6.5}
plot_pathways(pruned, top = 10L, sort_by = "divergence", min_count = 5L)
```

> **Reading the two heatmaps together.** This is the single best
> "interpretation" figure. Sorted by **count**, the heatmap is almost
> a vertical stripe on the Specification column — the monolithic
> default workflow, visually. Sorted by **KL** (with `min_count` to
> kill small-sample noise), the bright cells *move off* the
> Specification column: that lateral shift, side by side, is the entire
> thesis in one comparison — *common ≠ informative*. The `>` prefix
> marks modal-flip pathways; those rows are where history doesn't just
> sharpen but redirects (the Agreement/Positive-emotion family).

## 13.3 `plot_divergence()` — KL lollipop

```{r plot-divergence, fig.height = 5.5}
plot_divergence(pruned, top = 12L, min_count = 5L)
```

> **The numbers in plain language.** Each pathway carries a few numbers;
> read them as separate dials:
>
> | Column | In two words | What it tells you |
> |---|---|---|
> | `count` | *how common* | how often this history actually happened |
> | `divergence` (KL) | *history matters* | does remembering this step change the guess? |
> | `likely_next` | *best guess* | the single most likely next move |
> | `next_probability` | *how certain* | how confident that best guess is |
> | `changes_prediction` | *guess flips* | the best guess actually changed (orange) |
>
> The two that get conflated are **"history matters" (KL)** and **"how
> certain" (next_probability)** — they are independent. `Positive
> emotion` is high on *history matters* (1.13) yet low on *how certain*
> (39%): remembering it changes the guess a lot, but you are still not
> sure what comes next. `Agreement -> Positive emotion` is the opposite —
> lower on *history matters* (0.79) but very high on *how certain* (88%):
> the guess barely moves from the shorter context, yet it is a confident
> guess. Always read them as two dials, not one.

> **Reading it.** A ranked lollipop of per-pathway KL with orange
> points marking modal-flippers. With `min_count = 5L` the small-sample
> mirages from §3.4 are gone, so what remains is the honest divergence
> ranking: the orange (flipping) points at the top are the pathways
> that genuinely *change the prediction* — the ones to name in a
> results section. This is the figure that answers "so what is
> actually interesting in this data" in a single glance.

# 14. Uniform tidy-extract recap

```{r tidy-recap}
str(as.data.frame(pruned), give.attr = FALSE)
str(as.data.frame(boot),   give.attr = FALSE)
str(as.data.frame(cmp),    give.attr = FALSE)
str(as.data.frame(tg),     give.attr = FALSE)
```

> **Reading it.** Every result class — tree, bootstrap, comparison,
> tune — yields a tidy `data.frame` through the *same* `as.data.frame()`
> call, all sharing the canonical pathway columns where applicable.
> Practically: one extraction idiom, and results from different stages
> (`pruned`, `boot`, `cmp`) join on `pathway` without renaming. That is
> what made the cross-diagnostic argument for `Agreement -> Positive
> emotion` possible — the same key carried it through §3, §6 and §12.

# 15. What this dataset taught us (synthesis)

Pulling the thread through every section:

1. **Corpus is imbalanced** (§data): `Specification` ≈ ⅓ of events —
   so frequency is a misleading lens and modal predictions are
   trivially Specification.
2. **Open ≠ body** (§data): sessions *start* with `Instruct` (≈49%)
   but are *bodied* by `Specification`; a real behavioural asymmetry.
3. **Memory is short** (§1, §3.2, §6, §10): the unpruned tree was
   ~85% noise (G²/AIC), and cross-validation independently picks
   **order-2**. Depth-3/4 structure is overfit.
4. **The insight is in the rare codes** (§3.3, §4, §13.2): KL and
   `H_drop` concentrate on `Agreement` / `Positive emotion`, not the
   Specification machinery. *Common ≠ informative.*
5. **A reproducible micro-routine exists** (§3.4, §12): `Agreement ->
   Positive emotion` (`P ≈ 0.89`) survives top-KL, sharpness,
   *and* the bootstrap trust filter — the tutorial's firmest claim.
6. **Prompting matures over the course** (§11): Early = command +
   social loops; Late introduces `Context -> Reasoning`,
   `Reasoning -> Refine`; permutation-significant (p ≈ 0.01).

Each claim is anchored to a function whose output you can re-run —
which is the whole point of a pathway-centric, testable model.

# 16. Function reference index

| Function | Role | Section |
|---|---|---|
| `context_tree()` | fit a variable-depth tree (incl. native long event log) | §1 |
| `smooth_tree()` | re-smooth a fitted tree | §7 |
| `prune_tree()` | drop contexts that don't earn their depth | §6 |
| `tree_pathways()` | canonical 7-col tidy table | §3.1 |
| `common_pathways()` | top-n by frequency | §3.2 |
| `divergent_pathways()` | top-n by KL from parent | §3.3 |
| `sharp_pathways()` | top-n by predictive sharpness | §3.4 |
| `tree_dependence()` | per-context KL + entropy decomposition | §4 |
| `query_pathway()` | look up one pathway → distribution / prob | §5.1 |
| `pathway_exists()` | boolean: is this an exact node? | §5.2 |
| `subtree()` | restrict to descendants of a context | §5.3 |
| `predict.transitiontrees()` | next-state distribution for new histories | §8.1 |
| `perplexity()` | scalar predictive perplexity | §8.2 |
| `score_sequences()` | per-session held-out log-lik / perplexity | §8.2 |
| `score_positions()` | per-token log-lik + matched context | §8.2 |
| `logLik.transitiontrees()` | in-sample or held-out log-likelihood | §8.3 |
| `nobs.transitiontrees()` | observation count | §1, §8.3 |
| `generate_sequences()` | sample new sessions | §9.1 |
| `simulate.transitiontrees()` | S3 alias with `nsim` + `seed` | §9.2 |
| `tune_tree()` | k-fold CV grid search | §10 |
| `plot.transitiontrees_tune()` | perplexity surface | §10 |
| `tree_distance()` | scalar symmetric-KL between trees | §11.1 |
| `compare_trees()` | distance + permutation null | §11.2 |
| `plot.transitiontrees_comparison()` | null vs observed histogram | §11.3 |
| `bootstrap_pathways()` | stability + informativeness | §12 |
| `plot.transitiontrees_bootstrap()` | trust-quadrant forest plot | §12.3 |
| `plot_pathway_resamples()` | per-pathway resample histograms | §12.4 |
| `summary.transitiontrees_bootstrap()` | extract `$summary` | §12.1 |
| `plot.transitiontrees()` | dendrogram / horizontal / icicle / interactive | §13.1 |
| `plot_pathways()` | pathway × next-state heatmap | §13.2 |
| `plot_divergence()` | KL lollipop | §13.3 |
| `as.data.frame.*` | uniform tidy-extract | §2.2, §14 |

# 17. Session info

```{r session-info}
sessionInfo()
```
