---
title: "transitiontrees — A Tutorial Walkthrough"
subtitle: "Variable-order pathway modelling on student-engagement trajectories"
author: "transitiontrees"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    df_print: paged
    highlight: tango
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width = 9, fig.height = 5.2, dpi = 96
)
suppressPackageStartupMessages({
  devtools::load_all(here::here())
  library(ggplot2)
  data(trajectories, package = "Nestimate")
})
set.seed(2026)
```

# Why transitiontrees

Real trajectories — student engagement weeks, patient pathways,
clickstreams, play sequences — usually have memory that *varies by
context*. Some states predict the next state on their own; others
only when paired with a longer history. Fixed-order Markov chains
model "memory length 1, always" or "memory length 2, always".
**Variable-order context trees** let the data tell you, *for each
context*, how much history matters.

`transitiontrees` is the modern variable-order Markov / prediction-suffix-
tree (PST) toolkit for R. It ships the full **fit → prune → tune →
score → bootstrap → compare → simulate** pipeline. This tutorial
walks the whole surface on a real dataset.

------------------------------------------------------------------

# 1. The data

138 students, 15 weekly engagement states (Active / Average /
Disengaged); trailing NAs mark dropouts.

```{r data}
dim(trajectories)
table(unlist(trajectories), useNA = "ifany")
head(trajectories[, 1:8])
```

------------------------------------------------------------------

# 2. Fitting a pathway tree

The **pathway tree** is a variable-depth context tree. Each node
is a pathway (sequence of recent states) and carries a smoothed
conditional distribution over the next state.

```{r fit}
tree <- context_tree(
  trajectories,
  max_depth = 4L,    # longest pathway considered
  min_count      = 5L,    # min count for a context to get its own node
  smoothing = "floor"   # default; ymin = 0.001 used internally
)
tree
```

`summary()` prints a banner with a truncated overview of the
per-node table.

```{r summary}
summary(tree)
```

For the full tidy data.frame use `as.data.frame(tree)` (=
`pathways(tree)`). Section 4 uses it heavily.

------------------------------------------------------------------

# 3. Pruning

Prune nodes that don't earn their depth under a likelihood-ratio
test against the (k-1)-suffix.

```{r prune}
pruned <- prune_tree(tree, criterion = "G2", alpha = 0.05)
pruned
```

The pruned tree is the analysis object for several sections below.

------------------------------------------------------------------

# 4. The pathway-centric API

This is the consumer-facing view. Each row is a pathway with its
count, modal next state, sharpness (`prob_next`), KL vs. its
shorter-history parent, and a flag for whether the modal next
state *flips* between the long and short history.

```{r pathways}
common_pathways(pruned, top = 8)
```

```{r divergent}
divergent_pathways(pruned, top = 6)
```

```{r sharp}
sharp_pathways(pruned, top = 5)
```

```{r flips}
divergent_pathways(pruned, flips_only = TRUE, top = 8)
```

The `path_dependence()` table is the per-context KL diagnostic in
full — what the pruning rule was actually computing.

```{r pathdep}
head(tree_dependence(pruned), 12)
```

`plot_pathways()` and `plot_divergence()` give pathway-centric
views that complement (rather than replace) the tree plot.

```{r pathway-plot, fig.height = 4}
plot_pathways(pruned, n = 12)
```

```{r divergence-plot, fig.height = 4}
plot_divergence(pruned, n = 12)
```

------------------------------------------------------------------

# 5. Predictive evaluation

`transitiontrees` plays nicely with the `stats` model-comparison
toolchain. `logLik`, `AIC`, `BIC`, and a `perplexity()` metric
all work out of the box.

```{r logLik}
logLik(pruned)
AIC(pruned)
BIC(pruned)
perplexity(pruned)
```

A perplexity of ~`r round(perplexity(pruned), 2)` on a 3-state
alphabet says: *the model is as predictive as a uniform
distribution over `r round(perplexity(pruned), 2)` outcomes*. The
naïve baseline (fully uniform) is 3; perfect prediction is 1.

### 5.1 Holding out data

Held-out perplexity is the canonical PST evaluation metric. We
split the 138 students into a 70 / 30 train-test partition.

```{r holdout}
n <- nrow(trajectories)
test_idx  <- sample.int(n, size = round(0.3 * n))
train_idx <- setdiff(seq_len(n), test_idx)

tr_train <- context_tree(trajectories[train_idx, ],
                         max_depth = 4L, min_count = 5L)
tr_train <- prune_tree(tr_train, criterion = "G2", alpha = 0.05)

perplexity(tr_train, newdata = trajectories[test_idx, ])
```

Per-sequence and per-position scores:

```{r score-seq}
head(score_sequences(tr_train,
                     newdata = trajectories[test_idx, ]), 8)
```

```{r score-pos}
head(score_positions(tr_train,
                     newdata = trajectories[test_idx, ]), 8)
```

------------------------------------------------------------------

# 6. Smoothing schemes

Five smoothing schemes are implemented; `floor` is the default,
the others interpolate with the parent's smoothed distribution
along the suffix path.

```{r smoothing}
methods <- c("floor", "laplace", "kneser_ney",
             "witten_bell", "jelinek_mercer")
res <- vapply(methods, function(m) {
  tr <- context_tree(trajectories[train_idx, ],
                     max_depth = 4L, min_count = 5L,
                     smoothing = m)
  perplexity(tr, newdata = trajectories[test_idx, ])
}, numeric(1))
data.frame(smoothing = methods,
           held_out_perplexity = round(res, 4))
```

Smoothing matters for held-out scoring. With a small alphabet and
plenty of data, the differences are modest; on richer alphabets
or shorter sequences the gap can be much larger.

`smooth_tree()` re-smooths a fitted tree without refitting:

```{r resmooth}
tr_kn <- smooth_tree(pruned,
                          list("kneser_ney", discount = 0.5))
perplexity(tr_kn)
```

------------------------------------------------------------------

# 7. Cross-validated tuning

`tune_tree()` runs k-fold CV at the **sequence level** over a
grid of `(max_depth, nmin, smoothing, prune)` and returns a
ranked data.frame with `attr(., "best")` pointing at the
minimum-perplexity configuration.

```{r tune}
tg <- tune_tree(
  trajectories,
  max_depth = 1L:4L,
  min_count      = c(5L, 10L),
  smoothing = c("floor", "kneser_ney"),
  prune     = c(FALSE, TRUE),
  folds     = 5L,
  seed      = 42L
)
head(tg, 8)
```

```{r tune-best}
attr(tg, "best")
```

Built-in plot method:

```{r tune-plot, fig.height = 5}
plot(tg)
```

The chosen configuration is the model the CV procedure says will
predict best on unseen students — refit at this configuration on
the full data for downstream use.

------------------------------------------------------------------

# 8. Bootstrap pathway uncertainty

`prune()` answers "which pathways pass a parametric criterion
*in this dataset?*" The bootstrap answers "which pathways
*reproducibly* pass?" — a much stricter test.

`bootstrap_pathways()` is methodologically built on Saqr, Tikka &
López-Pernas (2025)
[**tna**](https://doi.org/10.1145/3706468.3706513), extending the
edge-level bootstrap framework to variable-depth pathways. It
operates **only on raw counts** — no `nmin` filter, no smoothing
inside the resampling loop, no extra parameters. Each iteration:

1. Resamples whole sequences with replacement.
2. Aggregates counts per depth from the precomputed
   per-sequence pair-index vectors.
3. Reads each original-tree pathway's count vector and computes
   empirical statistics (count, modal-next, prob_next, KL vs
   empirical parent, G² LR statistic vs empirical parent, modal
   flip flag).

Two complementary measures are reported per pathway:

- **`stability_rate`** — fraction of resamples where the chosen
  `stat` (default `count`) lies in
  `[0.75 × observed, 1.25 × observed]`. **Value
  reproducibility.** A pathway whose count drops to zero in a
  resample fails the band test automatically — the same
  mechanism `tna::bootstrap` uses for edge weights.
- **`informative_rate`** — fraction of resamples where the
  empirical likelihood-ratio statistic G² against the parent
  context exceeds the chi-square critical value at
  `alpha = 0.05`. **Reproducible significant divergence from
  the shorter-history baseline** — the bootstrap analogue of
  `prune(criterion = "G2")`.

A pathway is `stable` when `stability_rate ≥ 0.75` and
`informative` when `informative_rate ≥ 0.95` (strict, because at
α = 0.05 the chi-square test has a 5% Type-I rate; requiring 95%
of resamples to clear the threshold rules out finite-sample
chance crossings).

```{r boot}
boot <- bootstrap_pathways(tree, iter = 1000L, seed = 1L)
boot
```

The forest plot orders pathways by trustworthiness (stable +
informative first), shows the bootstrap CI on G² against the
chi-square critical value (dashed line), and colour-codes each
pathway by category:

```{r boot-plot, fig.height = 6}
plot(boot)
```

Raw bootstrap distribution of the chosen statistic per pathway:

```{r boot-resamples, fig.height = 4}
plot_pathway_resamples(boot, stat = "G2", top = 6L)
```

The full per-pathway view:

```{r boot-summary}
summary(boot)
```

### 8.1 Reading the trajectories result

The bootstrap finds **9 stable + informative** pathways out of
38 in the unpruned tree:

| depth | total | stable | informative | both |
|---:|---:|---:|---:|---:|
| 0 | 1 | 1 | NA | 0 |
| 1 | 3 | 3 | 3 | 3 |
| 2 | 9 | 8 | 5 | 5 |
| 3 | 25 | ~15 | 1 | 1 |

The depth-1 contexts (`Active`, `Average`, `Disengaged`) are
massively informative. Five of nine depth-2 contexts add
information beyond depth-1, including cohesive transitions like
`Active → Active` and `Disengaged → Disengaged`. **Only one
depth-3 context** (`Average → Active → Active`) clears both
bars; the remaining depth-3 nodes the unpruned tree fits are
either unstable counts (low frequency) or uninformative (their
depth-2 parent already carries the signal).

A pruned, defensible interpretation of trajectories therefore
has 9 pathways. Everything else is noise the unpruned tree
happened to fit.

### 8.2 Why two flags?

The two measures answer different questions:

- `stable` asks *"is the count value reproducible under
  resampling?"* — tna's framing. With enough data, counts
  reproduce regardless of whether the underlying structure is
  meaningful.
- `informative` asks *"does the LR test against parent
  reproducibly clear the chi-square threshold?"* — the test
  that distinguishes real signal from finite-sample noise.

A pathway worth interpreting is **both**. `stable` alone keeps
high-count noise pathways; `informative` alone could surface
low-count borderline pathways whose sample-level G² is high by
chance. The conjunction screens both failure modes.

### 8.3 Monte Carlo calibration

To verify the test isn't simply ratifying its own assumptions,
we ran a Monte Carlo evaluation across 600 simulated datasets —
200 datasets each from three regimes:

| regime | transition matrix | what it tests |
|---|---|---|
| STRONG | `P(modal) = 0.90` | high-power detection |
| WEAK | `P(modal) = 0.50` | moderate-effect detection |
| RANDOM | uniform `1/3` | Type-I rate |

For 1st-order Markov data, ground truth is unambiguous: depth-1
pathways are *truly informative* (their conditional differs from
the marginal); depth-2+ pathways are *truly NOT informative*
(under the Markov property, conditional on most recent state,
prior history adds nothing); for RANDOM, no pathway is
informative.

| regime | n | TP | FP | TN | FN | TPR | FPR | Precision | F1 |
|---|---:|---:|---:|---:|---:|---:|---:|---:|---:|
| STRONG | 2400 | 600 | 0 | 1800 | 0 | **1.000** | 0.000 | 1.000 | 1.000 |
| WEAK | 2400 | 600 | 2 | 1798 | 0 | **1.000** | 0.001 | 0.997 | 0.998 |
| RANDOM | 2400 | 0 | 1 | 2399 | 0 | — | 0.000 | — | — |

**0 false negatives** out of 1,200 truly-informative pathways.
**3 false positives** out of 6,000 truly-non-informative
pathways — well below the 5% nominal Type-I rate of the
underlying chi-square test, because we additionally require
95% resampling reproducibility. The empirical null G²
distribution at depth-2 across all three regimes (mean ≈ 2.6,
p95 ≈ 5.6) matches the chi-square_2 distribution (mean = 2,
p95 ≈ 6.0) — the test is properly calibrated.

------------------------------------------------------------------

# 9. Tree introspection

Three accessors expose the tree as a queryable object:

```{r query}
# Full conditional distribution at a specific pathway
query_pathway(pruned, c("Active", "Active"))

# Single-cell lookup
query_pathway(pruned, "Disengaged", next_state = "Disengaged")

# Membership test
pathway_exists(pruned, "Active -> Disengaged")
```

`subtree()` extracts a slice of the tree rooted at a given
pathway:

```{r subtree}
sub <- subtree(pruned, "Active")
sub
attr(sub, "local_root")
```

------------------------------------------------------------------

# 10. Comparing two cohorts

The differentiating PST primitive: **the tree IS the comparison
object**. To ask whether two student groups generate
significantly different pathway dynamics, fit a tree on each and
compare with a permutation test.

We split students by their early engagement (first 4 weeks) — a
contrived stratification, but illustrative.

```{r group-split}
early_active_count <- rowSums(trajectories[, 1:4] == "Active",
                               na.rm = TRUE)
group_a <- trajectories[early_active_count >= 3, ]
group_b <- trajectories[early_active_count <  3, ]
nrow(group_a); nrow(group_b)
```

```{r compare}
tr_a <- context_tree(group_a, max_depth = 3L, min_count = 5L)
tr_b <- context_tree(group_b, max_depth = 3L, min_count = 5L)

cmp <- compare_trees(tr_a, tr_b, iter = 100L, seed = 1L)
cmp
```

```{r compare-plot, fig.height = 4}
plot(cmp)
```

`tree_distance()` is the bare scalar without the permutation
test — useful for pairwise distance matrices over many cohorts.

```{r distance}
tree_distance(tr_a, tr_b)
```

------------------------------------------------------------------

# 11. Visualisation

Two ggplot-based plot styles, one dispatcher. No htmlwidget /
JavaScript dependencies.

### 11.1 Dendrogram (default)

Pure ggplot2. Node *size* = count; *fill* = KL vs. parent;
**thick ring** = nodes whose modal next state flips vs. their
shorter-history parent (the same modal-flip flag exposed by
`divergent_pathways()`). Labels show the most-recent state at
each node — read the full pathway by walking edges to root.

```{r plot-dendrogram, fig.height = 5}
plot(pruned)
```

### 11.2 Icicle (publication-ready partition)

`ggraph(layout = "partition")` renders a horizontal icicle. Tile
size is proportional to count; fill is KL vs. parent.

```{r plot-icicle, fig.height = 5}
plot(pruned, style = "icicle")
```

------------------------------------------------------------------

# 12. Generating new sequences

Every fitted tree is also a generative model: sample sequences
from its conditional distributions.

```{r generate}
gen <- generate_sequences(pruned, n = 5, length = 12)
gen
```

------------------------------------------------------------------

# 13. TraMineR `stslist` + per-sequence weights

`context_tree()` accepts TraMineR sequence objects natively and
honours their per-sequence weights.

```{r stslist, eval = requireNamespace("TraMineR", quietly = TRUE)}
suppressMessages(suppressWarnings({
  sts <- TraMineR::seqdef(trajectories)
}))
tr_sts <- context_tree(sts, max_depth = 3L, min_count = 5L)
identical(sort(tr_sts$alphabet), sort(tree$alphabet))
```

Per-sequence weights for the design-weighting case:

```{r weights, eval = requireNamespace("TraMineR", quietly = TRUE)}
w <- runif(nrow(trajectories), 0.5, 2)
tr_w <- context_tree(trajectories,
                     max_depth = 3L, min_count = 5L,
                     weights = w)
length(tr_w$nodes)
```

------------------------------------------------------------------

# 14. Reference / equivalence

`transitiontrees` was validated to **machine precision** against the
archived `PST` 0.94.1 (Gabadinho & Ritschard 2013) — counts
exact, smoothed probabilities max abs error 1.11 × 10⁻¹⁶ across
30 random configurations. The default `smoothing = "floor"` path
is byte-identical to the legacy implementation.

------------------------------------------------------------------

# Recap: the public surface

| Goal | Function |
|---|---|
| Fit | `context_tree()` |
| Prune | `prune()` |
| Pathway-centric inspection | `pathways()`, `common_pathways()`, `divergent_pathways()`, `sharp_pathways()`, `path_dependence()` |
| Score | `logLik()`, `AIC()`, `BIC()`, `perplexity()`, `score_sequences()`, `score_positions()` |
| Smooth | five schemes via `context_tree(smoothing = ...)`, `smooth_tree()` |
| Tune | `tune_tree()` |
| Bootstrap | `bootstrap_pathways()`, `plot_pathway_resamples()` |
| Query | `query_pathway()`, `subtree()`, `pathway_exists()` |
| Compare | `tree_distance()`, `compare_trees()` |
| Predict / Simulate | `predict()`, `generate_sequences()` |
| Visualise | `plot(style = c("dendrogram", "icicle"))`, `plot_pathways()`, `plot_divergence()` |

------------------------------------------------------------------

# References

- Begleiter, R., El-Yaniv, R., Yona, G. (2004). On prediction
  using variable-order Markov models. *JAIR*, **22**, 385–421.
- Bühlmann, P., Wyner, A. J. (1999). Variable length Markov
  chains. *Annals of Statistics*, **27**, 480–513.
- Gabadinho, A., Ritschard, G. (2013). Searching for typical
  life trajectories applied to childbirth histories. In:
  *Gendered Life Courses Between Standardization and
  Individualization*, 287–312.
- Phipson, B., Smyth, G. K. (2010). Permutation p-values should
  never be zero. *Statistical Applications in Genetics and
  Molecular Biology*, **9**(1), Article 39.
- Ron, D., Singer, Y., Tishby, N. (1996). The power of amnesia:
  learning probabilistic automata with variable memory length.
  *Machine Learning*, **25**, 117–149.
- **Saqr, M., Tikka, S., López-Pernas, S. (2025). Transition
  Network Analysis: a novel framework for modeling, visualizing,
  and identifying the temporal patterns of learners and learning
  processes.** *LAK '25*,
  doi:10.1145/3706468.3706513. — The edge-level bootstrap
  framework that transitiontrees's `bootstrap_pathways()` extends to
  variable-depth pathways.
- Willems, F. M. J., Shtarkov, Y. M., Tjalkens, T. J. (1995).
  The context-tree weighting method: basic properties.
  *IEEE Transactions on Information Theory*, **41**, 653–664.

------------------------------------------------------------------

# Session info

```{r session, echo = FALSE}
sessionInfo()
```
