---
title: "A worked `transitiontrees` example: AI-prompting move sequences"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width = 8, fig.height = 5, dpi = 110
)
library(here)
devtools::load_all(here::here())
library(ggplot2)
```

This is a short, end-to-end analysis of one real dataset — **not** a tour of
every function. The data are sequences of *rhetorical moves* students make
while prompting an LLM (`Order`, `Specify`, `Context`, `Reject`, `Request`,
`Approve`); one **prompt** is one sequence. The question throughout: *after a
given short history of moves, what does a student do next, and which contexts
overturn the corpus-wide default?*

# 1. Load and fit

`context_tree()` reads the long event log directly — give it the sequence id
(`actor`), the event column (`action`), and the within-sequence order
(`order`); it reshapes internally (no `Nestimate`/`prepare_input()` step
needed). Prompts are short, so the deepest context worth asking for is 3
moves; with only a few hundred prompts we keep `min_count` low.

```{r fit}
raw <- read.csv(
  "/Users/mohammedsaqr/Documents/Documents X/DocumentsX/TNA Work/TNA networks teem data/Seqdatalonged.csv",
  stringsAsFactors = FALSE)

tree <- context_tree(
  raw,
  actor     = "prompt_id",   # one prompt = one sequence
  action    = "name",        # the move
  order     = "value",       # move order within a prompt
  max_depth = 3L,            # longest context the tree may hold
  min_count = 3L             # a context needs >= 3 occurrences to get a node
)
tree
```

# 2. Inspect the model

The print banner gives depth, node count, and the corpus size. `summary()`
adds the per-depth breakdown, and `model_fit()` bundles the goodness-of-fit
scalars (`logLik`, `df`, `nobs`, `AIC`, `BIC`, `perplexity`) into one row.

```{r summaries}
summary(tree)
model_fit(tree)
```

The in-sample **perplexity** of `r round(perplexity(tree), 2)` says the fitted
tree is, on average, as uncertain as a fair choice among that many moves —
well below the 6-move uniform baseline, so history is genuinely informative.

# 3. Pathway tables

The pathway-centric API returns plain `data.frame`s in one canonical schema
(`pathway | depth | count | likely_next | next_probability | divergence |
changes_prediction`), so every view joins cleanly. Three named verbs each fix
a useful sort:

```{r tables}
common_pathways(tree, top = 8)      # by frequency: the backbone
divergent_pathways(tree, top = 8)   # by KL from the shorter history
sharp_pathways(tree, top = 8)       # by how peaked the next-move prediction is
```

`common_pathways()` is the backbone (what students mostly do).
`divergent_pathways()` surfaces the contexts where a longer history *changes*
the prediction — `changes_prediction = TRUE` flags a modal-move flip. Those
are the interesting rows: places where knowing the previous move(s) overturns
the corpus default.

# 4. Per-context diagnostics

`tree_dependence()` quantifies, per context, how much the history sharpens the
next-move distribution — `entropy_drop` is the reduction in predictive entropy
(bits) the context buys over its shorter suffix.

```{r dependence}
dep <- tree_dependence(tree)
head(dep[order(-dep$entropy_drop),
         c("pathway", "depth", "count", "entropy_drop",
           "likely_before", "likely_next", "changes_prediction")], 8)
```

# 5. Prune to the reliable tree

A context that survives fitting can still be over-fit. `prune_tree()` collapses
contexts whose extra depth does not buy a statistically significant gain over
their parent (default: a likelihood-ratio \eqn{G^2} test at `alpha = 0.05`).

```{r prune}
pruned <- prune_tree(tree, criterion = "G2", alpha = 0.05)
pruned
```

Pruning takes the model from `r n_nodes(tree)` contexts down to
`r n_nodes(pruned)` — the deepest, thinnest contexts that could not justify
their own parameters are folded back into their parents.

# 6. Bootstrap reliability

A pathway can look informative in one dataset by chance. `bootstrap_pathways()`
resamples whole prompts and asks which pathways are *reproducibly* stable
(count holds within a tolerance band) **and** informative (\eqn{G^2} clears the
\eqn{\chi^2} bar in most resamples). A pathway worth a claim is **both**.

```{r bootstrap}
boot <- bootstrap_pathways(pruned, iter = 200L, stat = "count", seed = 1L)
boot
```

```{r bootstrap-summary}
bs <- summary(boot)
bs[bs$stable & bs$informative,
   c("pathway", "depth", "count", "likely_next", "next_probability",
     "stability_rate", "informative_rate")]
```

These — the single-move backbone plus the handful of depth-2 rules that clear
the reproducibility bar — are the pathways you can state as findings.

# 7. Plots

A few of the package's visual outputs on this tree.

**The tree.** Node size scales with count, edge width with transition flow;
internal-node labels sit below their circles.

```{r plot-tree, fig.height=6}
plot(pruned, style = "horizontal")
```

**Pathway heatmap.** Each row is a context, each column a next move; the cell
is the predicted probability, the modal cell highlighted.

```{r plot-pathways, fig.height=6}
plot_pathways(tree, top = 14)
```

**Divergence lollipop.** Per-context KL from the shorter history; orange marks
contexts whose modal next move flips.

```{r plot-divergence, fig.height=6}
plot_divergence(tree, top = 12)
```

**Bootstrap forest.** Per-pathway \eqn{G^2} with bootstrap CIs; the dashed line
is the \eqn{\chi^2} critical value — pathways whose interval sits entirely to
its right are reproducibly informative.

```{r plot-bootstrap, fig.height=6}
plot(boot)
```

# 8. Predicting the next move

Finally, the point of the model: given a history, predict the next move and its
probability.

```{r predict}
predict(tree, c("Order", "Context"), type = "class")       # most likely next
predict(tree, c("Order", "Context"), type = "prob")        # full distribution
```
