---
title: "transitiontrees: A Complete, Interpreted Tutorial on the Grammar of AI Prompts"
subtitle: "Every function explained and every result read, on a real log of student prompt-writing moves"
author: "transitiontrees"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
      smooth_scroll: true
    code_folding: show
    df_print: paged
    highlight: tango
    fig_caption: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 9,
  fig.height = 5.5,
  dpi        = 96,
  comment    = "#>"
)
suppressPackageStartupMessages({
  devtools::load_all(here::here())
  library(ggplot2)
})
set.seed(20260602)
```

# How to read this tutorial

`transitiontrees` fits a **variable-order prediction suffix tree** (Ron, Singer
& Tishby 1996) and exposes a **pathway-centric API**. A fixed-order
Markov chain assumes memory is the *same length everywhere* ("the last
move predicts the next", or "the last two do"). That is rarely true of
behaviour. A variable-order tree lets the **data decide, per context,
how much history matters** — some prompting moves are predictable from
the single previous move, others only make sense given a longer run-up,
and most longer histories add nothing and should be pruned away.

This document does two things at once:

1. **Walks every exported function** against a real **log of how
   students compose AI prompts** — each prompt decomposed into its
   ordered *rhetorical moves* (`Order`, `Context`, `Specify`, …).
2. **Interprets every result.** After each table, test, and plot
   there is a *Reading* block: what the numbers actually say about how
   these students build a prompt, what the method is doing under the
   hood, and — just as important — *what you should not over-read*.

A running storyline emerges and is worth stating up front so the
sections connect rather than feel like a feature list:

> **The arc.** A prompt almost always *opens* with a command
> (`Order`) but its *body* is carried by `Specify` and `Context`. The
> single firmest rule in the corpus is grammatical: **after a
> `Context` move, students `Specify`** — the great majority of the time.
> The memory in this behaviour is **short** — prompts are only a few
> moves long, so cross-validation says order-2/3 is the ceiling and
> there is no deep structure to find. And the grammar **shifts over the
> course**: early assignments chain everything into `Specify`; later
> ones reorganise — a difference that is *statistically real*, whereas
> the same prompts split by **grade** show *no* difference at all. Every
> section below adds one piece of evidence to that arc.

Conventions:

- **Pathway** = a sequence of observed moves in arrow notation,
  `"A -> B -> C"` (leftmost = *oldest*).
- **Start** = the empty pathway (no history); `"(start)"` in tables,
  sentinel `"<root>"` internally. Its prediction is just the marginal
  "what move comes next on average".
- All public tables are plain `data.frame`s; the four result classes
  (`transitiontrees`, `transitiontrees_bootstrap`, `transitiontrees_comparison`,
  `transitiontrees_tune`) all support `as.data.frame()`.

# The dataset

The raw file is in **long format**: one row per *move*, carrying the
prompt it belongs to (`prompt_id`), the move's order within that prompt
(`value`), the move label (`name`), and a set of prompt-level covariates
(the student `id.x`, the `assignment`, the `llm` used, the graded
`performance`).

```{r data-load}
library(Nestimate)
raw <- read.csv(
  "/Users/mohammedsaqr/Documents/Documents X/DocumentsX/TNA Work/TNA networks teem data/Seqdatalonged.csv",
  stringsAsFactors = FALSE)
dim(raw)
str(raw)
sort(table(raw$name), decreasing = TRUE)
```

> **Reading the corpus.** Six rhetorical moves, and they are unevenly
> used (the counts are in the table above). `Order` and `Specify`
> together are the clear majority of every move made; `Approve` is
> almost absent. This imbalance is the single most important fact about
> the data and it echoes through every later result: a model that just
> guesses "Order" or "Specify" will look deceptively good, so the
> interesting question is never "what is the modal next move" (usually
> one of those two) but "*which contexts overturn that default*". Keep
> that lens for the rest of the tutorial.

The modelling unit is **one prompt** — `prompt_id` groups a prompt's
moves into one ordered sequence. Rather than reshape by hand, we let
`Nestimate::build_network()` do it: `actor = "prompt_id"` makes **each
prompt one sequence**, `order = "value"` puts its moves in order. The
returned `netobject` carries everything `transitiontrees` needs — `$data` is
the wide, trailing-NA-padded character frame (one row per prompt),
`$nodes$label` is the canonical alphabet (the move set), and
`$metadata` is row-aligned to `$data` with the prompt-level covariates
we use for the cohort comparison in §11.

```{r build-network}
Model <- build_network(raw, actor = "prompt_id", action = "name",
                       order = "value", method = "relative")
class(Model)
dim(Model$data)
head(Model$data)                        # trailing NAs = end-of-prompt
Model$nodes$label                       # the alphabet
```

```{r netobject-shape}
AI_data  <- Model$data                  # wide prompt frame, used throughout
alphabet <- Model$nodes$label           # the 6-move symbol space
length(unique(Model$metadata$id.x))             # distinct students
summary(rowSums(!is.na(AI_data)))               # prompt lengths
sort(Model$initial, decreasing = TRUE)          # OPENING move (netobject slot)
```

> **Reading the shape — and the first real finding.** Compare the
> *opening* move with the corpus frequencies above (what students
> *mostly do*). They disagree sharply: a prompt **opens with `Order`**
> far more often than `Order`'s one-third share of all moves — it is by
> far the commonest opening (see the printed opening distribution). That
> is a genuine behavioural signature: students *open* a prompt with a
> command ("build X", "make Y"), then spend the body **specifying and
> contextualising** what they actually want. We have not fitted anything
> yet and the data already has structure. Prompts are also **very
> short** — the printed `summary()` of prompt lengths shows a low median
> and a short ceiling. That brevity is *why* pruning and
> cross-validation will matter: a depth-3 context simply cannot have
> much evidence behind it, because only a handful of prompts run that
> long.

# 1. Fitting a context tree: `context_tree()`

`context_tree()` is the constructor. It returns an S3 `transitiontrees`: a
list of nodes keyed by pathway string, an edges table, the alphabet,
and the hyperparameters. The sibling-package integration: pass the
**Nestimate `netobject` directly** — it unwraps `$data` and uses
`$nodes$label` as the alphabet, so the network and the tree share one
symbol space. Because the deepest prompt is 4 moves, the deepest
*context* worth asking for is 3 (it predicts the 4th move), so we cap
`max_depth = 3`; with only a few hundred short prompts we keep `min_count`
low so depth-2/3 contexts can form at all.

```{r fit}
tree <- context_tree(
  Model,               # a Nestimate netobject, read natively
  max_depth = 3L,      # longest context the tree may represent
  min_count = 3L,      # a context needs >= 3 occurrences to get its own node
  smoothing = "floor"  # default: floor MLE at ymin=0.001 and renormalise
)
tree
```

> **Reading the fit.** Three numbers matter in that banner.
> **`r n_nodes(tree)` nodes** — far fewer than the AI-session trees you
> may have seen, precisely because the prompts are short: there are only
> so many distinct 2- and 3-move histories that occur >= 3 times.
> **`r tree$n_seq` sequences / `r nobs(tree)` observations** — every
> prompt with >= 1 move is kept. With *no* history the best guess is
> `Order` at `r round(query_pathway(tree, "(start)")["Order"], 2)`,
> barely ahead of `Specify`
> (`r round(query_pathway(tree, "(start)")["Specify"], 2)`); that
> near-tie is the baseline every deeper context has to *beat*. Hold that
> thought — §3 and §4 are the machinery for separating the rules that
> beat it from the spines that merely repeat.

The two things you actually reach for — how many contexts the tree
holds, and the root marginal — both have clean accessors, no
list-spelunking required:

```{r fit-internals}
n_nodes(tree)                    # how many contexts the tree holds
query_pathway(tree, "(start)")    # the root marginal = the null model
```

> **Reading the root marginal.** `query_pathway(tree, "(start)")` is the
> next-move distribution with *no* history (printed above): `Order` and
> `Specify` nearly tied at the top, `Context` next, the rest trailing.
> This vector is the **null model** for the entire analysis —
> "divergence from parent", G²
> pruning, and bootstrap "informativeness" all ultimately measure *how
> far a context's prediction departs from a chain of these marginals*.

## 1.1 Input shapes accepted

The wide matrix above is one of several equivalent front doors. The
same prompts can arrive as a ragged list, a TraMineR `stslist`, or — to
connect with the `mohsaqr` network ecosystem — a Nestimate
`netobject` / cograph network object built with
`session = "prompt_id"`.

We fit from the `netobject` above, but `context_tree()` accepts several
equivalent front doors — all numerically identical:

```{r fit-inputs, eval = FALSE}
context_tree(Model)                        # Nestimate netobject (this doc)
context_tree(Model$data)                   # the wide character frame
context_tree(TraMineR::seqdef(Model$data)) # TraMineR stslist
context_tree(cograph::as_cograph(Model))   # cograph network object
```

> **Reading the input rule.** The input shape is a convenience, never a
> numerical fork — every route above gives the same tree. The Nestimate
> route is the ecosystem handoff: `build_network(actor = "prompt_id",
> order = "value")` makes **each prompt one sequence**, hands `transitiontrees`
> the wide `$data` frame and `$nodes$label` as the alphabet, so the
> network and the tree share one symbol space. The hard boundary, by
> design: a **pure graph** object with no `$data` sequences — an
> aggregated transition network — is *rejected*, because the original
> move sequences cannot be recovered from edge weights. Numeric matrices
> are likewise rejected (they mean "transition matrix" elsewhere).
> *transitiontrees fits on sequences, never on aggregated transitions.*

## 1.2 The five smoothing schemes

Smoothing decides what probability an *unseen* next-move gets. With a
skewed 6-move alphabet and very short prompts, many context × move
cells are zero, so this choice is not cosmetic.

`compare_smoothing()` fits the tree under every scheme — holding
`max_depth`/`min_count` fixed — and returns the comparison in one call:

```{r fit-smoothing-grid}
compare_smoothing(Model, max_depth = 3L, min_count = 3L)
```

> **Reading the smoothing grid.** Two things. First, `n_nodes` is
> **identical across all five schemes** — smoothing changes
> *probabilities*, never *which contexts exist*; topology is set by
> `min_count`, not by the smoother. Second, the in-sample perplexities are
> close but ordered: plain `floor` looks best *in-sample* and `laplace`
> worst, because add-one `laplace` bleeds the most mass onto unseen
> moves — exactly the wrong move when two moves (`Order`/`Specify`)
> dominate. But *in-sample* perplexity rewards memorisation; do **not**
> pick a smoother here. §10's cross-validation reverses this ranking
> (`kneser_ney` wins out-of-sample) and that is the verdict that
> counts. The lesson is methodological: never select a hyperparameter
> on the data you fit on.

# 2. Inspecting the tree

## 2.1 `summary()` — tidy node table + scalars

```{r summary}
summary(tree)
```

> **Reading the summary.** The summary table is the canonical 7-column
> schema (`pathway, depth, count, likely_next, next_probability,
> divergence, changes_prediction`), sorted `(depth, -count)` so you read
> top-down from the root. The depth-1
> rows are the six single-move contexts — the backbone of everything.
> Glance at their `next_probability` in the printed table: `Context` is
> the most predictable (it points hard at `-> Specify`), `Order` and
> `Reject` middling, `Specify` itself the flattest (it can be followed
> by almost anything).
> `changes_prediction = TRUE` on these depth-1 rows means each move
> predicts a *different* next-move than the root's `Order` default —
> they carry real conditional information, which §3.3 quantifies.

## 2.2 `as.data.frame(tree)` — uniform tidy-extract

```{r as-data-frame}
df <- as.data.frame(tree)
head(df, 6)
all.equal(df, tree_pathways(tree))
```

> **Reading it.** `TRUE` — `as.data.frame(tree)` *is*
> `tree_pathways(tree)`; one source of truth, no schema drift
> between the console object and the tidy table you would put in a
> manuscript.

## 2.3 Model-fit scalars

`model_fit()` bundles all the standard fit scalars into one tidy row:

```{r model-fit-scalars}
model_fit(tree)
```

> **Reading the scalars.** Perplexity is the interpretable one:
> "effective number of equally-likely next moves". The uniform baseline
> is `r length(alphabet)` (six moves, no knowledge); the fitted tree is
> `r round(model_fit(tree)$perplexity, 2)` — knowing the recent moves
> collapses six possibilities to under three. That is a large, real
> effect, but it is *in-sample* and the depth-3 tree is mildly
> over-parameterised, so read it as an upper-optimistic bound. AIC/BIC
> exist to *penalise* node count; §10's cross-validation gives the
> honest, out-of-sample figure.

# 3. The pathway-centric API

Five exports return pathways as tidy data.frames sharing one schema:
`pathway | depth | count | likely_next | next_probability | divergence | changes_prediction`.

## 3.1 `tree_pathways()` — every node, sorted

```{r transitiontrees-pathways}
pw <- tree_pathways(tree)
nrow(pw); head(pw)
```

> **Reading it.** Default sort is `count` descending, so the top rows
> (printed above) are the corpus's *highways*: the root, then the single
> moves `Order` and `Context`, then `Order -> Context`. That last one is
> interesting already — a large share of prompts issue a command and
> then add context, and that 2-move history points almost
> deterministically onward. Frequency surfaces the *workflow*; the next
> two functions surface what is *informative*.

Schema is stable on empty results, and other sorts/filters work:

```{r transitiontrees-pathways-empty}
str(tree_pathways(tree, min_count = 1e6))   # 0 rows, still 7 typed cols
```

```{r transitiontrees-pathways-sort}
head(tree_pathways(tree, sort_by = "divergence"), 4)
head(tree_pathways(tree, sort_by = "depth", decreasing = FALSE), 4)
```

> **Reading it.** The empty case still returns the 7 typed columns — so
> downstream `rbind`/joins never break on a thresholded-away result.
> Re-sorting by `divergence` previews §3.3: the high-divergence rows are *different*
> pathways from the high-count rows — the contexts that *matter* are
> not the ones that are *common*.

## 3.2 `common_pathways()` — top-n by frequency

```{r common-pathways}
common_pathways(tree, top = 8)
```

> **Reading it.** The "what do students mostly do" view: prompts run on
> `Order` and `Context` openings that resolve into `Specify`. Note
> `Order -> Context -> Specify` and `Order -> Context` both point hard
> at `Specify` (see their `next_probability` above) — the dominant
> *constructive* pattern (command -> contextualise -> specify). Useful as
> a description of the common workflow; for the *interesting* behaviour
> you need divergence, not frequency.

```{r common-pathways-depth}
common_pathways(tree, top = 5, depth = 2L)
```

> **Reading it.** The top order-2 contexts are all `*-> Specify` or
> `Order -> Context`: the second move is highly predictable once you
> know the first. That the *frequent* 2-move histories are so
> deterministic is itself evidence for short, strong, local grammar —
> §10 makes the trade-off formal.

## 3.3 `divergent_pathways()` — top-n by KL from parent

```{r divergent-pathways}
divergent_pathways(tree, top = 8)
```

> **Reading it — the heart of the analysis.** KL here is, per context,
> *how much the prediction changes when you add one more step of
> history*, in nats. Read this table in two layers. The very top rows
> have enormous `divergence` **but a `count` of only 3** — a 2-move
> history seen three times that happened to resolve one way. Those are
> small-sample mirages; §3.4 and §12 exist to disarm them. The rows that
> *also* carry weight are the depth-1 moves — `Specify`, `Context` and
> `Order` — each with high `count` *and* substantial `divergence`. These
> are the well-supported redirections: each move pulls the prediction
> off the root's `Order` default toward a *specific* successor.
> `Context` is the standout — common *and* sharply predictive.

```{r divergent-pathways-flips}
divergent_pathways(tree, flips_only = TRUE, top = 8)
```

> **Reading the flippers.** Restricting to `changes_prediction = TRUE` keeps
> pathways where the longer history *overturns which move is most
> likely*, not merely sharpens it. Essentially the whole high-KL list
> survives — these divergences are genuine redirections. If this table
> had come back **empty**, the correct conclusion would have been "the
> tree only sharpens, never redirects" — a real, reportable negative.
> It is not empty, so these are the pathways to investigate, prune
> *toward*, and bootstrap in §12 — once we have filtered the count-3
> noise out.

## 3.4 `sharp_pathways()` — top-n by predictive sharpness

```{r sharp-pathways}
sharp_pathways(tree, top = 8)
```

> **Reading it — and a deliberate trap.** Sorted by `next_probability`
> descending, the top rows hit near-1.0 probabilities and look like
> near-deterministic rules. **Do not believe the low-`count` ones.** A
> 3-move context seen a handful of times that always continued the same
> way is exactly what a tree manufactures by chance. The rows that *are*
> trustworthy carry real support — `Order -> Context -> Specify` and its
> parent `Order -> Context`, both with high `count` in the table above.
> Same constructive routine, now with enough prompts behind it to take
> seriously. The lesson this table teaches is *why* §6 and §12 exist:
> sharpness without support is noise, and the package gives you formal
> tools to tell them apart rather than eyeballing `next_probability`.

# 4. Per-context diagnostics: `tree_dependence()`

This is the full information-theoretic decomposition that
`prune_tree(criterion = "KL")` thresholds: per context its
`divergence`, its entropy `entropy`, its parent's `entropy_before`, the
resolved uncertainty `entropy_drop = entropy_before - entropy`, and the
modal-flip flag (`changes_prediction`).

```{r path-dependence}
tree_dependence(tree, sort_by = "entropy_drop", top = 8)
```

> **Reading the decomposition.** `entropy_drop` answers "how much
> next-move uncertainty does this history *remove*?" **`Context` tops
> the table**: its parent (the root) is very uncertain over the six
> moves, and conditioning on a single `Context` move almost determines
> the next one (low `entropy`). It also *flips* the modal move from
> `Order` to `Specify`. That is the most valuable kind of context — it
> both sharpens and redirects, with a high `count` of support. The
> low-`count` rows below it show large `entropy_drop` too, but read them
> as candidates, not conclusions.

```{r path-dependence-base}
head(tree_dependence(tree, base = exp(1)), 4)   # nats instead of bits
```

> **Reading it.** Identical ordering, values rescaled by `ln 2` — base
> is presentation only. Use nats if you are reconciling against a
> likelihood (also in nats); bits if you are talking to people ("how
> many yes/no questions of uncertainty removed").

# 5. Specific-pathway lookup

The examples below use two concrete contexts from this alphabet: a
single move, and a two-move history written both ways `query_pathway()`
accepts — an arrow-string and a character vector.

```{r pick-contexts}
d1  <- "Context"               # a depth-1 context
d2  <- "Order -> Context"      # a depth-2 context, arrow form
d2v <- c("Order", "Context")   # the same context, vector form
```

## 5.1 `query_pathway()` — look up a single pathway

```{r query-pathway}
query_pathway(tree, d2)    # arrow-string form
query_pathway(tree, d2v)   # same context, vector form
```

> **Reading it.** Both call shapes (arrow-string and character vector)
> return the same conditional distribution over the six moves for the
> most-common depth-2 context, `Order -> Context`. Read it as a
> probabilistic rule: after a command then a context move, the next
> move is `Specify` with probability
> `r round(query_pathway(tree, d2)["Specify"], 2)` — the constructive
> routine made explicit.

```{r query-pathway-pinned}
query_pathway(tree, d1, next_state = "Specify")
```

> **Reading it.** Pinning `next_state` returns a single number:
> `P(Specify | last move = Context)` =
> `r round(query_pathway(tree, d1, next_state = "Specify"), 2)`. This is the primitive you
> use to answer targeted questions ("after a context move, how likely
> is a specify move?") without pulling the whole vector.

```{r query-pathway-exact}
deep_ctx <- c("Reject", "Approve", "Context")   # a 3-deep context unlikely to be a node
query_pathway(tree, deep_ctx)                # default: longest-suffix backoff
query_pathway(tree, deep_ctx, exact = TRUE)  # strict: all NA if not a node
```

> **Reading the backoff.** This contrast is the model's whole
> philosophy in two lines. By default an unseen 3-deep context **backs
> off to its longest matching suffix** (here `Context`) — exactly what
> makes a *variable*-order model robust: it never refuses to predict,
> it just uses as much history as it has evidence for. `exact = TRUE`
> returns all-`NA` for the same query, because that precise 3-gram was
> never a node. Use the default for prediction; use `exact = TRUE` only
> when auditing *which* contexts literally exist in the tree.

## 5.2 `pathway_exists()` — boolean check

```{r pathway-exists}
pathway_exists(tree, d1)
pathway_exists(tree, d2)
pathway_exists(tree, "nonsense -> path")
```

> **Reading it.** `TRUE/TRUE/FALSE`. The third is `FALSE` not because
> prediction failed but because that literal context is not a node —
> the membership test does *not* do suffix backoff, by design. It
> answers "did the tree learn this exact context", the right question
> for structure audits and test scaffolding.

## 5.3 `subtree()` — restrict to descendants

```{r subtree}
sub <- subtree(tree, d1)         # prints "subtree of: Context" in its banner
sub
head(tree_pathways(sub), 4)
```

> **Reading it.** You get a new, self-contained `transitiontrees` rooted at
> `Context` containing only that move and everything growing out of it,
> with the global arrow-form names preserved and the local root shown
> in its banner. Use this to drill into one motif — "everything that
> happens after a context move" — and run the *same* pathway API on the
> slice.

# 6. Pruning: `prune_tree()`

Pruning removes contexts that **don't earn their depth**: bottom-up, a
leaf whose evidence doesn't beat its parent under the chosen criterion
is dropped, the parent re-tested, and so on. The root is never pruned.

`compare_pruning()` applies each criterion — holding `alpha`/`threshold`
fixed — and reports how hard each one trims, in one call:

```{r prune-all}
compare_pruning(tree)
```

> **Reading the criteria.** The unpruned tree had `r n_nodes(tree)`
> nodes (the `n_nodes` column above is each criterion's surviving size).
> The likelihood-ratio rules agree tightly: **`G2`, `AIC` and `BIC` all
> cut it to about the same smaller tree**, while `KL` at the lenient
> 0.005-nat threshold keeps everything (any tiny divergence clears the
> bar — a knob, not a verdict; tighten `threshold` to make it bite). The
> contrast with a deep AI-session tree is instructive: there G²
> typically removes the large majority of nodes because long contexts
> overgrow wildly; **here it removes far less, because the prompts are
> so short there was little overgrowth to begin with.** The depth-3
> contexts that exist are mostly the low-`count` ones, and those are
> exactly what G²/AIC/BIC strip. **Practical guidance: use `G2` (or
> `AIC`); report the reduction, because "the unjustified contexts were
> the rare deep ones" is itself the short-memory finding.**

```{r prune-strict}
pr_strict <- prune_tree(tree, criterion = "G2", alpha = 0.001)
pr_strict   # banner reports node count + "pruned by: G2  alpha = 0.001"
```

> **Reading it.** Dropping `alpha` 0.05 → 0.001 raises the χ² bar, so
> even fewer contexts survive. The banner now reports the criterion and
> `alpha`, and those settings *travel with the tree* so
> `compare_trees()` re-applies the identical prune to every
> permutation refit (a correctness detail in §11).

```{r prune-pick}
pruned <- prune_tree(tree, criterion = "G2", alpha = 0.05)
n_nodes(pruned)
```

> **Reading it.** This `r n_nodes(pruned)`-node `pruned` tree is the **defensible
> model** and the working object for the rest of the tutorial:
> prediction, comparison, bootstrap and plots all use it, because the
> dropped contexts were the unsupported deep ones.

# 7. Re-smoothing a fitted tree: `smooth_tree()`

Smoothing is applied once at fit time, but `smooth_tree()` re-smooths
a fitted tree under a new scheme **without re-counting**:

```{r smooth-one}
pruned_kn <- smooth_tree(pruned, "kneser_ney")     # one scheme
c(original = n_nodes(pruned), re_smoothed = n_nodes(pruned_kn))  # topology unchanged
```

Handed a **fitted tree**, `compare_smoothing()` sweeps every scheme by
re-smoothing (not refitting), so the whole §1.2 comparison works on the
pruned model in one call:

```{r smooth-transitiontrees}
compare_smoothing(pruned)
```

> **Reading it.** Counts and topology are frozen — `n_nodes` is constant
> down the table — and only the `prob` vectors change, so perplexity
> moves a little while the *tree* is identical. (Contrast §1.2, where
> `compare_smoothing()` was given raw data and *refit* under each
> scheme.) This is the right tool for a smoothing sweep on an
> already-fit, already-pruned model — you are tuning how unseen moves
> are handled, not refitting structure.

# 8. Prediction surface

## 8.1 `predict.transitiontrees()`

Schema-stable rule: **container inputs (list/data.frame/matrix) always
return a matrix**; a bare character vector returns a named vector.

```{r predict-matrix}
predict(pruned, newdata = list(d2v, c("Order"), c("Reject", "Request")))
```

```{r predict-named-vec}
predict(pruned, newdata = d2v)
```

```{r predict-class}
predict(pruned, newdata = list(d2v, c("Order","Context"), c("Reject")),
        type = "class")
```

> **Reading it.** The matrix form (one row per history, one column per
> move) is for batch scoring; the named-vector shortcut is for
> interactive one-offs. `type = "class"` collapses each row to its
> modal move — and on this data those will lean toward `Specify` after
> any `Context`-bearing history and `Order` at the root, the predictable
> consequence of the grammar. When an exact context is missing from the
> pruned tree, prediction backs off to the longest matching suffix
> (same philosophy as §5.1), which is why this compact pruned tree still
> scores arbitrary new prompts.

## 8.2 `perplexity()`, `score_sequences()`, `score_positions()`

The scoring surface measures how well the model predicts a corpus. The
numbers here score the prompts the tree was fit on, so they are the
*optimistic* bound; the honest out-of-sample figure is the
cross-validation in §10 — no hand-rolled train/test split required.

```{r perplexity}
perplexity(pruned)
```

> **Reading it.** Perplexity is the "effective number of equally-likely
> next moves". The uniform baseline is `r length(alphabet)` (six moves,
> no knowledge); the pruned tree scores
> `r round(perplexity(pruned), 2)`, so knowing the recent history
> collapses six possibilities to under three. This is in-sample and
> therefore optimistic; §10 gives the cross-validated figure to report.

`score_sequences()` breaks that one number down to one row per prompt:

```{r score-sequences}
ss <- score_sequences(pruned, newdata = AI_data)
ss
```

> **Reading it.** One row per prompt with its own perplexity. Most
> prompts follow the common grammar and are predicted well; a tail is
> genuinely surprising — the worst prompt scores
> `r round(max(ss$perplexity), 1)` against a median of
> `r round(median(ss$perplexity), 1)`. Those high-perplexity prompts are
> where unusual rhetorical strategies live, worth qualitative follow-up.

`score_positions(worst =)` zooms to the individual moves the model
predicts worst:

```{r score-positions}
score_positions(pruned, newdata = AI_data, worst = 5)
```

> **Reading the worst predictions.** Each row is one observed move with
> the `matched_context` that drove its prediction, ranked by
> `predicted_prob` ascending — *where the model was most blindsided*:
> rare moves landing where the dominant `Order`/`Specify` machinery
> expected its usual successors. Diagnostic gold for which transitions
> more data would need to capture.

## 8.3 `model_fit()` — every fit scalar at once

`model_fit()` bundles the goodness-of-fit scalars into one tidy row, so
log-likelihood, the parameter count, AIC, BIC and perplexity come from a
single call:

```{r model-fit-pruned}
model_fit(pruned)
```

> **Reading it.** Per-observation log-likelihood (`logLik / nobs`) is
> just `-log(perplexity)` on the information scale that AIC/BIC tooling
> expects. AIC and BIC penalise the `r n_nodes(pruned)`-node count, so
> they are the right scalars to weigh two candidate trees fit on the
> same data.

# 9. Sequence generation

## 9.1 `generate_sequences()`

```{r generate}
gen <- generate_sequences(pruned, n = 5L, length = 4L)
gen
```

> **Reading it.** The tree is generative: each row is a synthetic
> prompt sampled by walking the conditional distributions. They look
> plausibly like real prompts — `Order`-opened, resolving into
> `Specify`/`Context` — a sanity check that the model captured the
> gross grammar. Generation is also how you build null behavioural
> corpora or stress-test a downstream pipeline. (Note `length = 4`
> matches the real prompt ceiling; asking for longer just keeps
> sampling from the same short-context rules.)

## 9.2 `simulate.transitiontrees()`

```{r simulate}
sim <- simulate(pruned, nsim = 5L, seed = 42L, length = 4L)
sim
```

> **Reading it.** The R-standard `simulate()` generic wrapping
> `generate_sequences()` with `nsim` + a `seed` for reproducibility —
> same draws as seeding `generate_sequences()` by hand. Use this form
> so your simulation slots into the usual `stats` workflow.

# 10. Cross-validated tuning: `tune_tree()`

k-fold CV over a hyperparameter grid; returns a `transitiontrees_tune`
data.frame sorted by held-out perplexity, with the winner on
`attr(., "best")`.

```{r tune}
tg <- tune_tree(AI_data, max_depth = 1L:3L, min_count = c(3L, 5L),
                    smoothing = c("floor","kneser_ney"),
                    prune = c(FALSE, TRUE), folds = 5L, seed = 1L)
head(tg, 8); attr(tg, "best")
```

> **Reading the verdict — this is where the arc lands.** The
> minimum-perplexity configuration is printed as `attr(tg, "best")`
> above — on this run, **`r attr(tg, "best")$smoothing`** at
> **`max_depth = r attr(tg, "best")$max_depth`** — but read the *whole
> table*, not just the winner. The perplexities are nearly flat: deeper
> contexts barely beat shallower ones, and the unpruned and pruned
> series sit almost on top of each other. Three conclusions follow. The
> out-of-sample smoothing winner **need not match the in-sample ranking
> from §1.2** — which is exactly why that earlier ranking had to be
> ignored: never select a smoother on the data you fit on. **Depth buys
> almost nothing** beyond order 2 — the quantitative form of "prompts
> have short memory". And **a winning `prune = FALSE` is no contradiction
> of §6**: the tree is already so compact (`r n_nodes(tree)` nodes) that
> there is little overgrowth for pruning to remove, so CV is indifferent
> to it — unlike a deep AI-session tree where pruning is decisive. The
> practitioner takeaway: a *shallow* model is the right production
> choice, and you would not bother shipping anything deeper.

```{r tune-plot, fig.height = 6}
plot(tg)
```

> **Reading the surface.** Perplexity (y) against the grid; the orange
> star is the winner. The shape is as informative as the point: the
> curves are essentially flat across `max_depth`, so there is no hidden
> deeper optimum — extra depth is wasted. A flat CV curve like this
> *is* the picture of a short-memory process; if prompt-writing had
> long-range structure you would see perplexity keep falling with
> depth, and it does not.

# 11. Two-cohort comparison

`compare_trees()` asks whether two cohorts generate *significantly
different* pathway dynamics: a count-weighted symmetric-KL distance plus
a permutation null from shuffling prompt → cohort labels. The natural
question for a *course* log is whether the grammar of a prompt **matures
over time**, so we split on `assignment`: **early** (1–2) vs **late**
(4–5).

We don't subset by hand: `context_tree(..., group =)` fits **one tree
per cohort** in a single call (over a shared alphabet), returning a
`transitiontrees_group`. `prune_tree()` prunes the whole group, and
`compare_trees()` takes the 2-group result directly. We fit at
`max_depth = 2` because each cohort is smaller.

```{r cohort-split}
stage   <- ifelse(Model$metadata$assignment %in% c("assignment1", "assignment2"),
                  "early", "late")
cohorts <- prune_tree(
  context_tree(Model, max_depth = 2L, min_count = 3L, group = stage),
  criterion = "G2", alpha = 0.05)
cohorts
tree_early <- cohorts$early; tree_late <- cohorts$late
```

## 11.1 `tree_distance()`

```{r transitiontrees-distance}
tree_distance(tree_early, tree_late)
tree_distance(tree_early, tree_late, symmetric = FALSE)
```

> **Reading it.** A single count-weighted symmetric-KL number
> (`r round(tree_distance(tree_early, tree_late), 3)`) —
> meaningful only against a null (next), but note `symmetric = FALSE`
> gives a *different* value: `D(early‖late) ≠ D(late‖early)`, because a
> pathway frequent early but absent late contributes differently than
> the reverse. Use the asymmetric form when you specifically care about
> "what does late do that early can't account for".

## 11.2 `compare_trees()`

```{r compare-transitiontreess}
cmp <- compare_trees(cohorts, iter = 200L, seed = 1L)  # 2-group -> pairwise
class(cmp); cmp
```

> **Reading the test — a real result.** Observed distance
> **`r round(cmp$pdist, 3)`**, permutation-null mean
> **`r round(mean(cmp$null_dist), 3)`**, **p =
> `r round(cmp$p_value, 3)`**. The observed separation sits well above
> what random relabelling produces, and clears the 0.05 bar:
> **the grammar of a prompt changes materially between early and late
> assignments.** The breakdown table below says *how*: early prompts
> funnel almost everything into a `Specify` move, while
> command-then-context (`Order -> Context`) and acknowledgement
> (`Approve`) show up far more in the late cohort. The reading:
> early-course prompting is a blunt "say it, then specify it" reflex;
> later prompting reorganises into command-then-contextualise and starts
> using acknowledgement moves. Students don't just prompt more; they
> prompt *with a different grammar* as they gain expertise.

```{r compare-transitiontreess-df}
head(as.data.frame(cmp), 8)
```

> **Reading the breakdown.** `count_a`/`count_b` are the raw cohort
> counts; the asymmetry is stark (some pathways are almost entirely
> early or almost entirely late — compare the two count columns above).
> `divergence_ab`/`divergence_ba`/`divergence_sym` quantify the
> Kullback-Leibler divergence each way. A row with a large count gap
> *and* large `divergence_sym` genuinely distinguishes the cohorts —
> those are the rows to
> quote, not the small-count ones whose KL is large only because a
> near-zero denominator inflates it.

```{r compare-transitiontreess-contrast}
## The same machinery on a split that does NOT separate: grade.
## group = "performance" reads the cohort straight from $metadata.
by_grade <- prune_tree(
  context_tree(Model, max_depth = 2L, min_count = 3L, group = "performance"),
  criterion = "G2", alpha = 0.05)
cmp_perf <- compare_trees(by_grade, iter = 200L, seed = 1L)
c(early_vs_late_p = cmp$p_value, high_vs_low_p = cmp_perf$p_value)
```

> **Reading the contrast — why the negative result matters.** Run the
> *identical* test on high- vs low-`performance` prompts and the result
> is **p = `r round(cmp_perf$p_value, 2)`** — set against the early/late
> **p = `r round(cmp$p_value, 3)`**, the grade split does not clear the
> 0.05 bar. This is the honest other half of the finding: the grammar of
> a prompt is shaped by **where in the course** it was written, but
> **not** by the grade it earned. A package that only ever returned
> "significant" would be useless; the value here is that it cleanly
> *fails to reject* when the structure isn't there. Report both: the
> maturation effect is real precisely because the grade effect, tested
> the same way, is not.

## 11.3 `plot.transitiontrees_comparison()`

```{r compare-transitiontreess-plot, fig.height = 4.5}
plot(cmp)
```

> **Reading the histogram.** The grey distribution is the
> label-shuffled null; the orange line is the observed distance. It
> sits in the right tail, *outside* the bulk of the null — that visual
> separation **is** the p = `r round(cmp$p_value, 3)`. If the line had landed inside the
> histogram (as it would for the grade split above), the Early/Late
> difference would have been indistinguishable from random relabelling
> and none of the interpretation above would be licensed. It does not,
> so it is.

# 12. Bootstrap reliability: `bootstrap_pathways()`

`prune_tree()` asked "which pathways pass a criterion *in this one
dataset*?". The bootstrap asks the stricter question: "which pass
*reproducibly* under resampling?" Per pathway: **`stability`** (its
count is reproducible within a tolerance band) and **`informative_rate`**
(its G² vs parent reproducibly clears χ²). A pathway worth trusting is
**both**.

```{r bootstrap}
boot <- bootstrap_pathways(pruned, iter = 200L, stat = "count",
          seed = 1L, keep_resamples = TRUE)
boot
```

> **Reading the quadrants — the trust filter.** Of the
> `r n_nodes(pruned)` pruned pathways,
> `r sum(summary(boot)$stable)` are **stable**,
> `r sum(summary(boot)$informative)` **informative**, and
> `r sum(summary(boot)$stable & summary(boot)$informative)` **both**.
> That last number is the one to internalise: only a handful of pathways
> in this entire corpus are simultaneously reproducible and predictively
> distinctive. The "both" set is five single-move pathways plus three
> depth-2 rules (`Order -> Specify`, `Context -> Specify`, `Reject ->
> Order`). Everything else — including the low-`count` high-divergence
> rows that §3.3 dangled in front of us — is not safe to build a
> narrative on. The depth-1 backbone is rock solid (huge, certain G²);
> the surviving depth-2 rules are the interesting ones because they
> cleared a much higher bar.

## 12.1 The summary table

```{r bootstrap-summary}
bs <- summary(boot)
bs[, c("pathway","depth","count","likely_next","next_probability",
       "divergence","changes_prediction","stable","informative",
       "stability_rate","informative_rate")]
```

> **Reading it.** First 7 columns are the canonical schema (so this
> table joins cleanly to every other one); then the flags and bootstrap
> stats. The pathway to single out is **`Context -> Specify`**: across
> §3.3 (high KL with support), §3.4 (high sharpness with real support),
> §4 (top `entropy_drop`) and now here it survives as `stable` *and*
> `informative`. That convergence of evidence across four independent
> diagnostics is what lets you state it as a finding: *after a student
> adds context to a prompt, a specification move follows reproducibly.*
> That is the tutorial's strongest second-order claim, and it is
> earned, not eyeballed.

```{r bootstrap-symmetric-stats}
grep("^(mean|sd|ci)_", names(bs), value = TRUE)
```

> **Reading it.** Every tracked stat (`count`, `next_probability`, `divergence`, `G2`)
> carries a symmetric `mean/sd/ci_lo/ci_hi` quartet — so you can report
> a bootstrap CI for any pathway statistic, not just a point estimate.
> Quote `ci_*` bounds in a paper; never the bare point.

## 12.2 `keep_resamples = FALSE`

```{r bootstrap-no-resamples}
boot_lean <- bootstrap_pathways(pruned, iter = 200L, seed = 1L,
                                keep_resamples = FALSE)
c(full = object.size(boot), lean = object.size(boot_lean))
```

> **Reading it.** Same `iter`, a fraction of the memory — the raw
> `iter x pathway` resample matrices are dropped on the lean object. The
> summary table is unaffected; you only lose the ability to draw §12.4's
> histograms (the plotter errors helpfully if you try). Use this for big
> `iter` in production; keep the default when you still want the
> distributions.

## 12.3 `plot.transitiontrees_bootstrap()`

```{r bootstrap-plot, fig.height = 7}
plot(boot, top = 15)
```

> **Reading the forest plot.** Each row a pathway; bars are 95% CIs on
> the bootstrap G²; the dashed line is the χ² critical value. The rule
> is purely visual: **a bar entirely to the right of the dashed line is
> reproducibly informative**; a bar straddling it is not safe. Colour
> encodes the trust quadrant. The depth-1 moves have bars far to the
> right (huge, certain G²); the two surviving depth-2 rules sit closer
> to the line — real, but with the wider CIs that honest small-sample
> inference should show. This is the plot to put in a supplement: it
> shows reliability and effect size together.

## 12.4 `plot_pathway_resamples()`

```{r bootstrap-resamples-plot, fig.height = 5.5}
plot_pathway_resamples(boot, stat = "next_probability", top = 6L)
```

> **Reading the histograms.** One panel per top pathway, the full
> bootstrap distribution of `next_probability`. Shape matters: a tight,
> symmetric peak means the estimate is well-determined; a **bimodal or
> heavy-tailed** panel is the tell that the pathway is *carrier-driven*
> — a handful of prompts (or students) account for it, and dropping
> them in a resample collapses it. For any pathway you intend to claim,
> check its panel here is unimodal and tight before you trust the point
> estimate from §3.

# 13. Visualization

In every static style: **node fill** = last move of the pathway,
**node size** = context count, **edge thickness** = the child's count
(the volume of prompts flowing down that branch).

## 13.1a Horizontal phylogram (default)

```{r plot-default, fig.height = 8, fig.width = 10}
plot(pruned)
```

> **Reading it.** `(start)` on the left, depth rightward along smooth
> curved branches, every leaf labelled with its **full arrow-form
> pathway**. This is the default style and the layout to use in a paper
> when you need to *cite specific pathways inline* — the reader can
> trace `Context -> Specify` to its leaf and read its encodings without
> decoding a circular layout. The thick `Order` and `Context` branches
> are the corpus highways; the thin outer twigs are contexts the G²
> prune *kept on evidence, not volume*.

## 13.1b Radial dendrogram

```{r plot-radial, fig.height = 7}
plot(pruned, style = "dendrogram")
```

> **Reading the picture.** The eye should go to the thick edges near
> the centre: the high-volume `Order` and `Context` branches — the
> corpus highways. The thin outer twigs are the contexts the G² prune
> *kept on evidence, not volume*. The visual tension between a few fat
> central branches and a sparse fringe *is* the dataset's story in one
> image: most prompt-writing is a short command-then-specify reflex; the
> structure lives in the first two moves.

```{r plot-sized, fig.height = 7}
plot(pruned, style = "dendrogram",
     point_size_range = c(4, 14), edge_size_range = c(0.4, 4))
```

> **Reading it.** Identical topology, exaggerated size dynamic range —
> for slides/posters where the count contrast needs to read from the
> back of a room. Encodings unchanged; only the scales differ.

## 13.1c Circular icicle (sunburst)

```{r plot-icicle, fig.height = 7}
plot(pruned, style = "icicle")
```

> **Reading it.** A space-filling partition: arc *angular width* is
> proportional to count, so the dominant `Order` and `Specify` wedges
> visually swallow the ring — an immediate, honest depiction of the
> move imbalance that §The-dataset quantified. Inner ring = depth-1
> (full labels), outer rings abbreviated; arcs too thin to label are
> the rarer contexts. Best when you want to *show the imbalance itself*
> as the message.

## 13.1d Interactive network tree

```{r plot-interactive, fig.height = 6, fig.width = 9}
plot(pruned, style = "interactive")
```

> **Reading it.** A `visNetwork` widget carrying the *same* encoding as
> the static styles — **node size = context count, edge width = the
> child's count (flow)**, fill = last move — but draggable and zoomable,
> with a hover tooltip for the full pathway, count, modal next move and
> distribution. This is the *exploration* tool: the fat central `Order`
> and `Context` nodes and thick branches are the corpus highways, while
> the thin outer twigs are the evidence-kept contexts. Drag a node to
> pull a motif clear of the rest, or hover `Context -> Specify` to read
> its reliable rule directly.

## 13.2 `plot_pathways()` — pathway × next-move heatmap

```{r plot-pathways, fig.height = 6.5}
plot_pathways(pruned, top = 12L, sort_by = "count")
```

```{r plot-pathways-kl, fig.height = 6.5}
plot_pathways(pruned, top = 10L, sort_by = "divergence", min_count = 5L)
```

> **Reading the two heatmaps together.** This is the single best
> "interpretation" figure. Sorted by **count**, the heatmap concentrates
> on the `Order`/`Specify` columns — the monolithic default workflow,
> visually. Sorted by **KL** (with `min_count` to kill small-sample
> noise), the bright cells *move*: `Context`-bearing rows light up the
> `Specify` column and `Order` lights up off-diagonal. That shift, side
> by side, is the entire thesis in one comparison — *common ≠
> informative*. The `>` prefix marks modal-flip pathways: rows where
> history doesn't just sharpen but redirects.

## 13.3 `plot_divergence()` — KL lollipop

```{r plot-divergence, fig.height = 5.5}
plot_divergence(pruned, top = 12L, min_count = 5L)
```

> **The numbers in plain language.** Each pathway carries a few numbers;
> read them as separate dials:
>
> | Column | In two words | What it tells you |
> |---|---|---|
> | `count` | *how common* | how often this history actually happened |
> | `divergence` (KL) | *history matters* | does remembering this step change the guess? |
> | `likely_next` | *best guess* | the single most likely next move |
> | `next_probability` | *how certain* | how confident that best guess is |
> | `changes_prediction` | *guess flips* | the best guess actually changed (orange) |
>
> The two that get conflated are **"history matters" (KL)** and **"how
> certain" (next_probability)** — they are independent. `Specify` is high
> on *history matters* (1.11) yet low on *how certain* (38%): remembering
> it changes the guess a lot, but you are still not sure. `Context` is
> high on both. Always read them as two dials, not one.

> **Reading it.** A ranked lollipop of per-pathway KL with orange
> points marking modal-flippers. With `min_count = 5L` the count-3
> mirages from §3.3 are gone, so what remains is the honest divergence
> ranking: the orange (flipping) points at the top — `Context`,
> `Specify`, `Order`, `Context -> Specify` — are the contexts that
> genuinely *change the prediction*, the ones to name in a results
> section. This is the figure that answers "so what is actually
> interesting in this data" in a single glance.

> **What the ranking reveals.** Order-sensitivity concentrates on the
> directive moves — `Specify`, `Order`, and `Context` — which carry the
> largest divergences and the prediction-overturning flips. The process
> is organised around commanding rather than negotiating. These
> high-divergence contexts are almost all shallow, depth-1 histories:
> the single most recent directive governs what follows, not an
> accumulated multi-step trajectory. The deeper depth-2 contexts that
> would signal genuine deliberation are fewer, thinly supported, lower
> in divergence, and tend to confirm rather than redirect the default —
> there is little evidence of cumulative depth. Equally telling is what
> is absent: the evaluative moves of a negotiated exchange barely shape
> prediction. `Approve` is among the rarest moves, and where `Reject`
> appears it routes straight back into another directive
> (`Reject -> Specify → Order`, `Reject → Specify`). The pattern is
> one-directional and commanding: the most informative thing about any
> moment is the last directive given, and the sequence advances by
> issuing the next one rather than by evaluating or building on what
> came before.

# 14. Uniform tidy-extract recap

```{r tidy-recap}
str(as.data.frame(pruned), give.attr = FALSE)
str(as.data.frame(boot),   give.attr = FALSE)
str(as.data.frame(cmp),    give.attr = FALSE)
str(as.data.frame(tg),     give.attr = FALSE)
```

> **Reading it.** Every result class — tree, bootstrap, comparison,
> tune — yields a tidy `data.frame` through the *same* `as.data.frame()`
> call, all sharing the canonical pathway columns where applicable.
> Practically: one extraction idiom, and results from different stages
> join on `pathway` without renaming. That is what made the
> cross-diagnostic argument for `Context -> Specify` possible — the same
> key carried it through §3, §4, §6 and §12.

# 15. What this dataset taught us (synthesis)

Pulling the thread through every section:

1. **The move alphabet is imbalanced** (§data): `Order` + `Specify` are
   the clear majority of all moves — so frequency is a misleading lens
   and modal predictions are trivially one of those two.
2. **Open ≠ body** (§data): prompts *open* with `Order` far more than
   its overall share but are *bodied* by `Specify`/`Context`; a real
   behavioural asymmetry — command first, then refine.
3. **Memory is short** (§1, §6, §10): the prompts are only a few moves
   long, G²/AIC/BIC trim the rare deep contexts, and cross-validation
   shows depth-3 ≈ depth-2. There is no long-range structure to find.
4. **The insight is in the second move** (§3.3, §4, §13.2): divergence
   and `entropy_drop` concentrate on `Context`/`Specify` openings, not
   on length. *Common ≠ informative.*
5. **A reproducible micro-rule exists** (§3.4, §4, §12): `Context ->
   Specify` survives top-divergence, sharpness, top `entropy_drop`
   *and* the
   bootstrap trust filter — the tutorial's firmest second-order claim.
6. **The grammar matures over the course** (§11): early assignments
   funnel everything into `Specify`; late ones add command-then-context
   and acknowledgement; permutation-significant (p =
   `r round(cmp$p_value, 3)`) — while the same test on **grade** is *not*
   significant (p = `r round(cmp_perf$p_value, 2)`), which is what makes
   the maturation effect credible.

Each claim is anchored to a function whose output you can re-run —
which is the whole point of a pathway-centric, testable model.

# 16. Function reference index

| Function | Role | Section |
|---|---|---|
| `context_tree()` | fit a variable-depth tree (wide / list / `stslist` / netobject / tna) | §1 |
| `context_tree(group=)` | batch-fit one tree per group → `transitiontrees_group` | §11 |
| `print.transitiontrees_group()` / `as.data.frame.transitiontrees_group()` | inspect / tidy-extract a group | §11 |
| `smooth_tree()` | re-smooth a fitted tree (or group) | §7 |
| `compare_smoothing()` | compare all smoothing schemes (refit from data, or re-smooth a fitted tree) | §1.2, §7 |
| `prune_tree()` | drop contexts that don't earn their depth (or prune a whole group) | §6, §11 |
| `compare_pruning()` | one-call comparison of all pruning criteria | §6 |
| `tree_pathways()` | canonical 7-col tidy table | §3.1 |
| `common_pathways()` | top-n by frequency | §3.2 |
| `divergent_pathways()` | top-n by KL from parent | §3.3 |
| `sharp_pathways()` | top-n by predictive sharpness | §3.4 |
| `tree_dependence()` | per-context KL + entropy decomposition | §4 |
| `query_pathway()` | look up one pathway → distribution / prob | §5.1 |
| `pathway_exists()` | boolean: is this an exact node? | §5.2 |
| `subtree()` | restrict to descendants of a context | §5.3 |
| `predict.transitiontrees()` | next-move distribution for new histories | §8.1 |
| `perplexity()` | scalar predictive perplexity | §8.2 |
| `score_sequences()` | per-prompt log-lik / perplexity | §8.2 |
| `score_positions()` | per-move log-lik + matched context | §8.2 |
| `model_fit()` | one-call fit scalars (logLik/df/nobs/AIC/BIC/perplexity) | §2.3, §8.3 |
| `logLik.transitiontrees()` | in-sample or held-out log-likelihood | §8.3 |
| `nobs.transitiontrees()` | observation count | §1 |
| `n_nodes()` | context (node) count — accessor for `length(tree$nodes)` | §1, §5.3, §6 |
| `generate_sequences()` | sample new prompts | §9.1 |
| `simulate.transitiontrees()` | S3 alias with `nsim` + `seed` | §9.2 |
| `tune_tree()` | k-fold CV grid search | §10 |
| `plot.transitiontrees_tune()` | perplexity surface | §10 |
| `tree_distance()` | scalar symmetric-KL between trees | §11.1 |
| `compare_trees()` | distance + permutation null (accepts a 2-group `transitiontrees_group`) | §11.2 |
| `plot.transitiontrees_comparison()` | null vs observed histogram | §11.3 |
| `bootstrap_pathways()` | stability + informativeness | §12 |
| `plot.transitiontrees_bootstrap()` | trust-quadrant forest plot | §12.3 |
| `plot_pathway_resamples()` | per-pathway resample histograms | §12.4 |
| `summary.transitiontrees_bootstrap()` | extract `$summary` | §12.1 |
| `plot.transitiontrees()` | dendrogram / horizontal / icicle / interactive | §13.1 |
| `plot_pathways()` | pathway × next-move heatmap | §13.2 |
| `plot_divergence()` | KL lollipop | §13.3 |
| `as.data.frame.*` | uniform tidy-extract | §2.2, §14 |

# 17. Session info

```{r session-info}
sessionInfo()
```
