---
title: "transitiontrees on the trajectories data"
subtitle: "One analysis, its default output, and how to read it — nothing more"
author: "transitiontrees"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_depth: 2
    toc_float:
      collapsed: false
    df_print: paged
    highlight: tango
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 9,
  fig.height = 6,
  dpi        = 96,
  comment    = "#>"
)
suppressPackageStartupMessages({
  devtools::load_all(here::here())
  library(ggplot2)
})
```

This is a plain walk-through of `transitiontrees` on the built-in
`trajectories` data. Every step is a single **default call** — no
subsetting, no tuning, no clever code. Each section is one analysis, its
default output or plot, and a short reading.

# The data

```{r data}
data(trajectories)            # bundled with transitiontrees
dim(trajectories)
head(trajectories)
```

> One row per learner, one column per time-step, three engagement
> states (`Active`, `Average`, `Disengaged`). Trailing `NA`s mark the
> end of a trajectory. This wide frame is exactly what `context_tree()`
> consumes.

# Fit a context tree

```{r fit}
tree <- context_tree(trajectories)
tree
```

> The banner is the whole model at a glance: how many contexts it holds,
> the alphabet, how many sequences and observations it was fit on, and —
> at the bottom — the next state predicted from no history (the `(start)`
> row) followed by the tree of deeper contexts. With the defaults the
> tree grows to depth 5, so it is large and over-grown; the next step
> trims it.

# Prune the tree

```{r prune}
pruned <- prune_tree(tree)
pruned
```

> The default fit above grows all the way to depth 5 and keeps deep
> contexts the data barely supports (over-fitting). Pruning keeps a
> longer history only when it predicts better than its shorter version,
> so the tree gets much smaller — the banner now reads `pruned by: G2`.
> The rest of the vignette uses this pruned tree.

# Draw the tree

```{r plot-tree, fig.height = 8}
plot(pruned)
```

> The default picture: a left-to-right tree. **Node size** is how often a
> context occurs, **branch thickness** is the volume flowing down it, and
> **colour** is the most recent state. The thick branches are the common
> engagement patterns; the thin twigs are the rarer contexts pruning kept
> on evidence.

# Most common pathways

```{r common}
common_pathways(pruned)
```

> The trajectories the data simply contains the most copies of, ranked by
> count. `likely_next` and `next_probability` say what tends to come
> after each — the everyday flow of engagement.

# Most divergent pathways

```{r divergent}
divergent_pathways(pruned)
```

```{r plot-divergence}
plot_divergence(pruned)
```

> These are the contexts where knowing the longer history **changes the
> prediction most** (the `divergence` column, drawn as the lollipop).
> Orange points mark pathways where the extra history actually flips
> which state is most likely — the genuinely informative histories,
> as opposed to the merely common ones above.

# Sharpest pathways

```{r sharp}
sharp_pathways(pruned)
```

> Ranked by `next_probability`: the contexts whose next state is most
> nearly certain. High sharpness with a high `count` is a dependable
> rule; high sharpness on a small `count` is just chance, and the
> bootstrap section below separates the two.

# Per-context diagnostics

```{r dependence}
tree_dependence(pruned)
```

> The information-theoretic view of each context: its `divergence`, its
> `entropy` (how uncertain its next state is), the parent's
> `entropy_before`, and `entropy_drop` — how much uncertainty the context
> removes. A large `entropy_drop` with `changes_prediction = TRUE` is the
> most valuable kind of context: it both sharpens and redirects.

# Pathway × next-state heatmap

```{r heatmap, fig.height = 7}
plot_pathways(pruned)
```

> The same pathways as a grid: each row a context, each column a next
> state, each cell `P(next | context)`. The bright cells are the strong
> rules; a `>` prefix marks a context whose history flips the most likely
> next state.

# Goodness of fit

```{r fit-scalars}
model_fit(pruned)
```

> The standard scalars in one row. `perplexity` is the readable one — the
> effective number of equally-likely next states; against a uniform
> baseline of three states, the tree's value shows how much the recent
> history narrows things down. `AIC`/`BIC` penalise the node count and
> let you weigh two candidate trees.

# Reliability under resampling

```{r bootstrap}
boot <- bootstrap_pathways(pruned)
boot
```

```{r plot-bootstrap, fig.height = 7}
plot(boot)
```

> The bootstrap refits the tree on resampled learners and asks which
> pathways survive *reproducibly*. The forest plot shows each pathway's
> bootstrap interval against the significance line; a bar entirely to
> the right is reproducibly informative. The pathways that are both
> stable and informative are the ones worth building a claim on.

# Other views of the same tree

```{r plot-icicle, fig.height = 7}
plot(pruned, style = "icicle")
```

```{r plot-radial, fig.height = 7}
plot(pruned, style = "dendrogram")
```

> The same pruned tree as a sunburst and as a radial dendrogram — the
> encodings (size = count, fill = state, depth = rings or radius) are
> identical, only the layout changes. There is also
> `plot(pruned, style = "interactive")` for a draggable version.

# Session info

```{r session}
sessionInfo()
```
